---
title: "Read-based chronic wound virome analysis"
output: html_notebook
---

This notebook can be used to generate figures from the manuscript, using the tables included in the repo. Keep the notebook and tables in the same directory to avoid file path issues.


# Libraries
```{r}
# Load the required libraries
library(phyloseq)
library(genefilter)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(RColorBrewer)
library(viridis)
library(wesanderson)
library(grid)
library(gridExtra)
library(plotly)
library(scales)
library(dplyr)
library(cowplot)
set.seed(1)
```

# File Import & Reformatting
```{r}
# vOTU import
biom_file <- "ps2.virome.abundance.dryad.csv"
otu.csv <- read.csv(biom_file, row.names = 1, header = TRUE, check.names = FALSE)
ps2.otu.mat <- as.matrix(otu.csv, rownames=TRUE)
mode(ps2.otu.mat) <- "numeric"

# Import mapping file
map_file <- "ps2mapping.ps2vb.dryad.txt"

# Build phylo object
bmsd <- import_qiime_sample_data(map_file)
OTU <- otu_table(ps2.otu.mat, taxa_are_rows = TRUE)
SAM <- sample_data(bmsd)
ps2 <- merge_phyloseq(OTU, SAM)


# import and add taxonomy
tax_file <- "ps2.virome.taxtable.dryad.csv"
ps2.tax <- read.csv(tax_file, row.names = 1, header = TRUE, check.names = FALSE, na.strings=c("","NA"))
ps2.tax_table <- as.matrix(ps2.tax)
tax_table(ps2) <- ps2.tax_table
```

# Decontamination w/ Decontam
## preview library size
```{r}
library(decontam)
# preview library size
ps <- ps2
df <- as.data.frame(sample_data(ps))
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
raw.lib.size <- ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point() + ylim(1e+03, 5e+06) + scale_y_log10()
raw.lib.size
png("raw.lib.size.png", width = 6, height = 4, units = "in", res = 600)
raw.lib.size
dev.off()
```

## identify contaminants
```{r}
# Assign negative controls
sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Negative Control"
contamdf.prev <- isContaminant(ps, method="prevalence", neg="is.neg", 
                               threshold = 0.2)
table(contamdf.prev$contaminant)
```
## visualize contaminants
```{r}
ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "Negative Control", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "True Sample", ps.pa)
# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                      contaminant=contamdf.prev$contaminant)
decon.plot <- ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")

decon.plot

decon.plot$data

png("contam.prevalence.png", width = 6, height = 6, units = "in", res = 600)
decon.plot
dev.off()
```
# Begin pre-processing
## Subset tables
```{r}
## Sample subsetting
### Samples only
ps2.pf <- subset_samples(ps2, SampleType %in% c("Pre-Debridement","Post-Debridement","Skin"))
ps2.pf
#### Controls only
ps2.con <- subset_samples(ps2, SampleType %in% c("Negative Control", "Mock Community"))
ps2.con
```

## Filter contaminants
```{r}
library(tibble)
# Decontam contaminant list
contamdf.prev$otuid <- rownames(contamdf.prev)
contam.otus <-  contamdf.prev %>%
    filter(contaminant == "TRUE")
rownames(contam.otus) <- contam.otus$otuid
contam.otus

# add all ambiguous taxa from neg cons
flist <- filterfun(kOverA(1,1))
ps2.con.ambig <- filter_taxa(ps2.con, flist, TRUE)
ps2.con.ambig <- subset_taxa(ps2.con.ambig, TResolution == "Ambiguous")
ps2.con.ambig <- as.data.frame(tax_table(ps2.con.ambig))
ps2.con.ambig$otuid <- rownames(ps2.con.ambig)
contam.combined <- as.character(c(ps2.con.ambig$otuid, contam.otus$otuid))

# Print Decontam list to file
contam.list <- prune_taxa(contam.combined, ps2.pf)
contam.list
contam.table <- tax_table(contam.list)
write.csv(contam.table, "decontam.OTUs.csv")

# filter from sample object
deconned.taxa <- setdiff(taxa_names(ps2.pf), contam.combined)
## Filter the table to only include decontaminated OTU IDs
ps2.pf.fil <- prune_taxa(deconned.taxa, ps2.pf)
# compare
ps2.pf
ps2.pf.fil


# Manually remove additional contaminants
ps2.pf.decon <- subset_taxa(ps2.pf.fil, !(Species %in% c("Escherichia virus M13", "Escherichia virus DE3", "BeAn 58058 virus", "Escherichia virus T1", "Escherichia virus T4", "Escherichia virus T7", "Escherichia virus Lambda", "Human endogenous retrovirus K", "Vibrio virus CTXphi")))
ps2.pf.decon <- subset_taxa(ps2.pf.decon, !(Type %in% c("Escherichia phage", "Enterobacteria phage", "Enterobacteriaceae phage", "Shigella phage", "Xanthomonas phage", "Orthobunyavirus", "Betabaculovirus")))

# preview
ps2.pf.decon
```


```{r}
# deconned lib size
ps.dlibsize <- ps2.pf.decon
df <- as.data.frame(sample_data(ps.dlibsize))
df$LibrarySize <- sample_sums(ps.dlibsize)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
decon.lib.size <- ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point() + ylim(1e+03, 5e+06) + scale_y_log10()
decon.lib.size
png("decon.lib.size.png", width = 6, height = 4, units = "in", res = 600)
decon.lib.size
dev.off()

```

# Taxonomic agglommeration and pre-melting
```{r}
## Subset Ambiguous and Defined Species & Host/Types
ps2.sgd.ss <- subset_taxa(ps2.pf.decon, SResolution == "Defined")
ps2.tgd.ss <- subset_taxa(ps2.pf.decon, TResolution == "Defined")

## Agglommerate at the Species and Type levels
### Sample objects
ps2.sg <- tax_glom(ps2.pf.decon, "Species", NArm = FALSE)
ps2.tg <- tax_glom(ps2.pf.decon, "Type", NArm = FALSE)
ps2.sgd <- tax_glom(ps2.sgd.ss, "Species", NArm = FALSE)
ps2.tgd <- tax_glom(ps2.tgd.ss, "Type", NArm = FALSE)
ps2.tpsg <- tax_glom(ps2.tgd.ss, "Type_PS", NArm = FALSE)
### Control objects
ps2.con.sg <- tax_glom(ps2.con, "Species", NArm = FALSE)
ps2.con.tg <- tax_glom(ps2.con, "Type", NArm = FALSE)

## Pre-melt count objects
### Sample counts
ps2.pf.melt <- psmelt(ps2.pf.decon)
ps2.sg.melt <- psmelt(ps2.sg)
ps2.sgd.melt <- psmelt(ps2.sgd)
ps2.tg.melt <- psmelt(ps2.tg)
ps2.tgd.melt <- psmelt(ps2.tgd)
ps2.tpsg.melt <- psmelt(ps2.tpsg)
### Control Counts
ps2.con.melt <- psmelt(ps2.con)
ps2.con.sg.melt <- psmelt(ps2.con.sg)
ps2.con.tg.melt <- psmelt(ps2.con.tg)

# Count normalization
## Samples Overall
ps2.ra <- transform_sample_counts(ps2.pf.decon, function(x) (x / sum(x))*100 )
## Samples - Species Glommed
ps2.sgra <- transform_sample_counts(ps2.sg, function(x) (x / sum(x))*100 )
ps2.sgdra <- transform_sample_counts(ps2.sgd, function(x) (x / sum(x))*100 )
## Samples - Type Glommed
ps2.tgra <- transform_sample_counts(ps2.tg, function(x) (x / sum(x))*100 )
ps2.tgdra <- transform_sample_counts(ps2.tgd, function(x) (x / sum(x))*100 )
ps2.tpsgra <- transform_sample_counts(ps2.tpsg, function(x) (x / sum(x))*100 )

## Control count normalization
## Overall
ps2.con.ra <- transform_sample_counts(ps2.con, function(x) (x / sum(x))*100 )
## Species glommed
ps2.con.sgra <- transform_sample_counts(ps2.con.sg, function(x) (x / sum(x))*100 )
## Type glommed
ps2.con.tgra <- transform_sample_counts(ps2.con.tg, function(x) (x / sum(x))*100 )

## Pre-melting RA objects
### Sample relative abundance
ps2.ra.melt <- psmelt(ps2.ra)
ps2.sgra.melt <- psmelt(ps2.sgra)
ps2.tgra.melt <- psmelt(ps2.tgra)
ps2.sgdra.melt <- psmelt(ps2.sgdra)
ps2.tgra.melt <- psmelt(ps2.tgra)
ps2.tgdra.melt <- psmelt(ps2.tgdra)
ps2.tpsgra.melt <- psmelt(ps2.tpsgra)
### Control Relative abundance
ps2.conra.melt <- psmelt(ps2.con.ra)
ps2.con.sgra.melt <- psmelt(ps2.con.sgra)
ps2.con.tgra.melt <- psmelt(ps2.con.tgra)

```

# Collect sample counts after full decontamination
```{r}
library(tidyr)
viral.counts <- data.frame(Raw = sample_sums(ps2.pf))
viral.counts$Decontaminated <- sample_sums(ps2.pf.decon)
viral.counts$Defined <- sample_sums(ps2.tgd)
df.temp <- sample_data(ps2.pf)
viral.counts$Type <- df.temp$SampleType
df <- viral.counts
df$Index <- seq(nrow(df))
df

# deconned lib size
df_long <- df %>% 
  gather(key, value, -Index, Type)
df_long

decon.process.sum1 <- df %>% ggplot(aes(x=reorder(Index, -Raw))) + 
  geom_point(aes(y=Raw, color="Raw", shape=Type)) +
  geom_point(aes(y=Decontaminated, color="Decontaminated", shape=Type)) +
  geom_point(aes(y=Defined, color="Defined", shape=Type)) +
  ylim(1e+03, 5e+06) + scale_y_log10() +
  ylab("Viral read abundance") + xlab("Sample index") + theme_bw() +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "bottom") +
  labs(color="Decontamination step") +
  scale_shape_discrete(limits = c("Pre-Debridement", "Post-Debridement", "Skin")) +
  scale_color_discrete(limits = c("Raw", "Decontaminated", "Defined")) +
  guides(shape = guide_legend(ncol = 1, title.position = "top", title.hjust = 0.5)) +
  guides(color = guide_legend(ncol = 1, title.position = "top", title.hjust = 0.5))
decon.process.sum1


#decon.process.sum
png("decontamination.process.summary.png", width = 6, height = 6, units = "in", res = 600)
decon.process.sum1
dev.off()
```


# Control Analysis
## Viral Composition

```{r}
## for decontaminated control test
# filter from sample object
deconned.taxa.con <- setdiff(taxa_names(ps2.con.ra), contam.combined)
## Filter the table to only include decontaminated OTU IDs
ps2.con.decon <- prune_taxa(deconned.taxa, ps2.con.ra)


# Manually remove additional contaminants
ps2.con.decon <- subset_taxa(ps2.con.decon, !(Species %in% c("Escherichia virus M13", "Escherichia virus DE3", "BeAn 58058 virus", "Escherichia virus T1", "Escherichia virus T4", "Escherichia virus T7", "Escherichia virus Lambda", "Human endogenous retrovirus K", "Vibrio virus CTXphi")))
ps2.con.decon <- subset_taxa(ps2.con.decon, !(Type %in% c("Escherichia phage", "Enterobacteria phage", "Enterobacteriaceae phage", "Shigella phage", "Xanthomonas phage", "Orthobunyavirus", "Betabaculovirus")))

# compare
ps2.con
ps2.con.decon
```

```{r}
# TAXONOMIC BARCHART
# Filter by # of most abundant taxa
ps2.conbox <- subset_samples(ps2.con.sgra, SampleType == "Negative Control")
ps2.conbox <- prune_taxa(names(sort(taxa_sums(ps2.conbox),TRUE)[1:30]), ps2.conbox)

# Convert to relative abundance
ps2.conbox = transform_sample_counts(ps2.conbox, function(x) x / sum(x) )

#assign specific colors to specific Family 
set.seed(132)
getPalette <- colorRampPalette(brewer.pal(50, "Set1"))
ConFamList <- unique(tax_table(ps2.conbox)[,"Species"])
ConFamList <- sample(ConFamList)
ConFamPalette = getPalette(length(ConFamList))
names(ConFamPalette) = ConFamList

# make a boxplot
sp1 = plot_bar(ps2.conbox, fill="Species", x="SampleID") + theme(legend.position="bottom") + scale_fill_manual(values = ConFamPalette)
sp1 = sp1 + theme_bw() + theme(legend.position = "bottom",
        text = element_text(size=20),
        strip.text.y = element_text(size=24),
        axis.title = element_text(size=26),
        axis.text = element_text(size=22),
        legend.title = element_text(size=24)) +
        guides(fill = guide_legend(title = "Species", ncol=2)) + xlab("Sample ID")

sp1
# Print & save to file
png("controls.viralspecies.taxplot.png", width = 12, height = 14, units = "in", res = 600)
sp1
dev.off()
```
### Filter out contaminants and replot
```{r}
sp1$data <- filter(sp1$data, !(OTU %in% contam.combined))
sp1$data <- filter(sp1$data, !(Species %in% c("Escherichia virus M13", "Escherichia virus DE3", "BeAn 58058 virus", "Escherichia virus T1", "Escherichia virus T4", "Escherichia virus T7", "Escherichia virus Lambda", "Human endogenous retrovirus K", "Vibrio virus CTXphi")))
sp1$data <- filter(sp1$data, !(Type %in% c("Escherichia phage", "Enterobacteria phage", "Enterobacteriaceae phage", "Shigella phage")))


# plot
sp1 + ylim(0, 100)

# Print & save to file
png("controls.viralspecies.taxplot.DECONNED.png", width = 12, height = 14, units = "in", res = 600)
sp1 + ylim(0, 100)
dev.off()
```

## Control type glommed composition
```{r}
# TAXONOMIC BARCHART
# Filter by # of most abundant taxa
ps2.conbox <- ps2.con.tgra
ps2.conbox <- subset_samples(ps2.con.tgra, SampleType == "Negative Control")
ps2.conbox <- prune_taxa(names(sort(taxa_sums(ps2.conbox),TRUE)[1:15]), ps2.conbox)

# Convert to relative abundance
ps2.conbox = transform_sample_counts(ps2.conbox, function(x) x / sum(x) )

#assign specific colors to specific Family 
set.seed(132)
getPalette <- colorRampPalette(brewer.pal(50, "Set1"))
ConFamList <- unique(tax_table(ps2.conbox)[,"Type"])
ConFamList <- sample(ConFamList)
ConFamPalette = getPalette(length(ConFamList))
names(ConFamPalette) = ConFamList

# make a boxplot
sp1t = plot_bar(ps2.conbox, fill="Type", x="SampleID") + scale_fill_manual(values = ConFamPalette)
sp1t =  sp1t + theme_bw() + theme(legend.position = "bottom",
        text = element_text(size=20),
        strip.text.y = element_text(size=24),
        axis.title = element_text(size=26),
        axis.text = element_text(size=22),
        legend.title = element_text(size=24)) + xlab("Sample ID") +
        guides(fill = guide_legend(title = "Type", ncol=2))

sp1t

# Print & save to file
png("controls.viraltype.taxplot.png", width = 10, height = 10, units = "in", res = 600)
sp1t
dev.off()


library(cowplot)

png("negative.control.bars.png", width = 22, height = 16, units = "in", res = 1000)
plot_grid(sp1t, sp1, ncol=2, align="h")
dev.off()

```

### Filter contaminants and replot
```{r}
sp1t$data <- filter(sp1t$data, !(OTU %in% contam.combined))
#sp1$data <- filter(sp1$data, !(Species %in% c("Escherichia virus M13", "Escherichia virus DE3", "BeAn 58058 virus", "Escherichia virus T1", "Escherichia virus T4", "Escherichia virus T7", "Escherichia virus Lambda", "Human endogenous retrovirus K", "Vibrio virus CTXphi")))
sp1t$data <- filter(sp1t$data, !(Type %in% c("Escherichia phage", "Enterobacteria phage", "Enterobacteriaceae phage", "Shigella phage", "Poxvirus", "Vibrio phage")))

sp1t + ylim(0, 100) + xlab("Sample ID")
# Print & save to file
png("controls.viraltype.taxplot.DECONNED.png", width = 10, height = 10, units = "in", res = 600)
sp1t + ylim(0, 100)
dev.off()
```


# Viral control box and whiskers
```{r}
# remove escherichia phage

test.sp2 <- ps2.con.tgra.melt %>%
  group_by(Description, Type) %>%
  filter(Abundance > 0.1) %>%
  filter(SampleType == "Negative Control") %>%
ggplot(aes(x= reorder(Type, +Abundance), y = Abundance)) +
  geom_point(stat='identity') +
  geom_boxplot(fill="orange") +
  coord_flip() + 
  theme_bw() +
  theme(legend.position="none") +
  xlab("Type") + ylab("Relative Abundance (%)")

test.sp2

# Print & save to file
png("control.viral.boxwhisk.TYPE.png", width = 4.5, height = 6, units = "in", res = 600)
test.sp2
dev.off()
```
## Positive Control expected/observed
### prepare expected data
```{r}
# Enter mock data
mock.species <- c("Escherichia virus T7", "Escherichia virus T4", "Escherichia virus Lambda", "Escherichia virus M13", "Escherichia virus phiX174")
mock.mce = c(25.8, 54.55, 16.71, 1.95, 0.99)
mock.mcl = c(23, 62.35, 14.33, 0.21, 0.10)
# build into dataframe
pos.comp <- data.frame(mock.species, mock.mce, mock.mcl)
pos.comp$Sample[1:5] <- "Expected"
pos.comp
```
### prepare observed data and build table
```{r}
# prepare viral mock expected data
## Subset
poscon <- subset_samples(ps2.con.ra, SampleType == "Mock Community")
poscon <- subset_taxa(poscon, Species %in% mock.species)
poscon <- tax_glom(poscon, "Species")

## melt
posmelt <- psmelt(poscon)

## average the samples and filter to include only taxa in mock community
pos.avg <- posmelt %>% group_by(SampleID, Species) %>% summarise(sum = sum(Abundance))
obmce <- pos.avg %>% filter(SampleID == "MCE")
obmcl <- pos.avg %>% filter(SampleID == "MCL")

# build observed table
obtab <- data.frame(obmce$Species)
colnames(obtab)[1] <- "mock.species"
obtab$mock.mce <- obmce$sum
obtab$mock.mcl <- obmcl$sum
obtab$Sample[1:5] <- "Observed"

# combine expected and observed tables
pos.comp <- rbind(pos.comp, obtab)

### plot pos con data
colnames(pos.comp)[2] <- "Mock_1"
colnames(pos.comp)[3] <- "Mock_2"
pos.comp

# plot mock 1
pcp1 <- pos.comp %>% 
  group_by(Sample, mock.species) %>%
  ggplot(aes(x=Sample, y=Mock_1, fill = mock.species)) +
  geom_bar(stat="identity") +
  theme_bw() + ylab("Average relative abundance (%)") + 
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_fill_brewer(palette = "Set1", name = "Species") + ggtitle("Mock Community 1")
pcp1

pcp2 <- pos.comp %>% 
  group_by(Sample, mock.species) %>%
  ggplot(aes(x=Sample, y=Mock_2, fill = mock.species)) +
  geom_bar(stat="identity") +
  theme_bw() + ylab("Average relative abundance (%)") + 
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_fill_brewer(palette = "Set1", name = "Species") + ggtitle("Mock Community 2")
pcp2

poscon.grid <- plot_grid(pcp1, pcp2, align = "h", axis = "b", rel_widths = c(1, 1))

poslegend <- get_legend(
  # create some space to the left of the legend
  pcp1 + theme(legend.position = "bottom")
)



png("positive.control.comp.png", width = 10, height = 5, units = "in", res = 800)
plot_grid(poscon.grid, poslegend, ncol = 1, rel_heights = c(1, .1))
dev.off()

```


# Overall Viral Abundance Barchart by Host
```{r}
# Calculate average relative abundance of each host-glom, plot
p1 <- ps2.tgdra.melt %>%
  group_by(Type) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  ggplot(aes(x= reorder(Type, +mean.abundance), y = mean.abundance, fill = mean.abundance)) +
  geom_bar(stat='identity') +
  coord_flip() + 
  scale_fill_gradientn(colours = c('gray','darkolivegreen3','darkolivegreen4')) +
  #scale_fill_manual(values = wes_palette("Darjeeling1", type = "discrete")) +
theme_light() +
  theme(legend.position="none", text = element_text(size=15)) +
  xlab("Type") + ylab("Average relative abundance (%)")
p1

# Print & save to file
png("type.viral.relabund.summary.DEFINED.png", width = 6, height = 7, units = "in", res = 1000)
p1
dev.off()

p1s <- ps2.sgdra.melt %>%
  group_by(Species) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  ggplot(aes(x= reorder(Species, +mean.abundance), y = mean.abundance, fill = mean.abundance)) +
  geom_bar(stat='identity') +
  coord_flip() + 
  scale_fill_gradientn(colours = c('gray','magenta3','magenta4')) +
  #scale_fill_manual(values = wes_palette("Darjeeling1", type = "discrete")) +
  theme_light() +
  theme(legend.position="none", text = element_text(size=15)) +
  xlab("Species") + ylab("Average relative abundance (%)") #+ scale_y_log10()
p1s

# Print & save to file
png("viral.species.relabund.summary.DEFINED.png", width = 7, height = 7, units = "in", res = 1000)
p1s
dev.off()


png("overall.abundance.DECONNED.png", width = 8, height = 6, units = "in", res = 1000)
plot_grid(p1, p1s, ncol=2, align="h")
dev.off()
```


```{r}
## AVG & STDEV
unclass.stats <- ps2.pf.decon
tax_table(unclass.stats) <- tax_table(unclass.stats)[,7:6]
unclass.stats <- tax_glom(unclass.stats, "TResolution")
unclass.stats <- transform_sample_counts(unclass.stats, function(x) (x / sum(x))*100 )
unclass.stats.melt <- psmelt(unclass.stats)

avg.unclass <- unclass.stats.melt %>% group_by(TResolution) %>% summarise(mean = mean(Abundance))
sd.unclass <- unclass.stats.melt %>% group_by(TResolution) %>% summarise(sd = sd(Abundance))
avg.unclass
sd.unclass

all.sums <- data.frame(sample_sums(ps2.pf))
all.sums %>% summarise(mean = mean(sample_sums.ps2.pf.))
all.sums %>% summarise(sd = sd(sample_sums.ps2.pf.))

```

# Viral Taxonomic Stacked Barchart (Species Level)
## Filter to top 20 OTUs
```{r}
# filter to most abundant taxa
## top 30 species overall
ps2.gb.200 <- prune_taxa(names(sort(taxa_sums(ps2.sgdra),TRUE)[1:50]), ps2.sgra)
```
## Plot

```{r}
# Convert to relative abundance
ra.tf.ps2 = transform_sample_counts(ps2.gb.200, function(x) x / sum(x) )

# make a boxplot, sort by patient, stack pre-, post-, skin
p2g = plot_bar(ra.tf.ps2, fill="Species", facet_grid = "SampleType~.", x="Patient") #+ scale_fill_manual(values = genus.colors)
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
# facet order
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
p2g$data$SampleType <- as.character(p2g$data$SampleType)
p2g$data$SampleType <- factor(p2g$data$SampleType, levels=newSTorder)
# x-axis order
newPTorder = as.character(c(1:20))
p2g$data$Patient <- as.character(p2g$data$Patient)
p2g$data$Patient <- factor(p2g$data$Patient, levels=newPTorder)

p2gf = p2g + theme_bw() + 
        theme(legend.position = "bottom",
        text = element_text(size=20),
        strip.text.y = element_text(size=24),
        axis.title = element_text(size=26),
        axis.text = element_text(size=22),
        legend.title = element_text(size=24)) +
  guides(fill = guide_legend(title = "Species", ncol=3))
p2gf


# Print & save to file
png("viral.species.taxplot.wlegend.DEFINED.png", width = 15, height = 18, units = "in", res = 600)
p2gf
dev.off()

```

# Viral Taxonomic Boxplots (Type level)
```{r}
# filter to most abundant taxa
ps2.hgp <- ps2.tpsgra
ps2.hgp <- prune_taxa(names(sort(taxa_sums(ps2.hgp),TRUE)[1:30]), ps2.hgp)
```

## Plot

```{r}
# Convert to relative abundance
#ps2.hgp.ra = transform_sample_counts(ps2.hgp, function(x) x / sum(x) )

# make a boxplot, sort by patient, stack pre-, post-, skin
p2hg = plot_bar(ps2.hgp, fill="Type_PS", facet_grid = "SampleType~.", x="Patient") #+ scale_fill_manual(values = genus.colors)
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
# facet order
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
p2hg$data$SampleType <- as.character(p2hg$data$SampleType)
p2hg$data$SampleType <- factor(p2hg$data$SampleType, levels=newSTorder)
# x-axis order
newPTorder = as.character(c(1:20))
p2hg$data$Patient <- as.character(p2hg$data$Patient)
p2hg$data$Patient <- factor(p2hg$data$Patient, levels=newPTorder)

p2gf = p2hg + theme_bw() + 
        theme(legend.position = "bottom",
        text = element_text(size=20),
        strip.text.y = element_text(size=24),
        axis.title = element_text(size=26),
        axis.text = element_text(size=22),
        legend.title = element_text(size=24)) +
  guides(fill = guide_legend(title = "Type", nrow=10))
p2gf

# Print & save to file
png("viral.typePS.taxplot.wlegend.DEFINED.png", width = 12.5, height = 14, units = "in", res = 600)
p2gf
dev.off()
```


# Interactive Taxonomic Barchart
```{r}
# INTERACTIVE TAXPLOT
# Transform genus-glommed table to relative abundances, filter
ra.gg1pct.ps2 = filter_taxa(ps2.sgdra, function(x) mean(x) > 0.01, TRUE)

set.seed(132)
getPalette <- colorRampPalette(brewer.pal(26, "Set1"))
#getPalette <- colorRampPalette(brewer.pal(22, "Spectral"))
#getPalette <- colorRampPalette(brewer.pal(22, "Accent"))
GenList <- unique(tax_table(ra.gg1pct.ps2)[,"Type"])
GenList <- sample(GenList)
GenPalette = getPalette(length(GenList))
names(GenPalette) = GenList


# make a boxplot, sort by patient, stack pre-, post-, skin
p2gi = plot_bar(ra.gg1pct.ps2, fill="Type", facet_grid = "SampleType~.", x="Patient") #+ scale_fill_manual(values = GenPalette)
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
# facet order
p2gi$data$SampleType <- as.character(p2gi$data$SampleType)
p2gi$data$SampleType <- factor(p2gi$data$SampleType, levels=newSTorder)
# x-axis order
p2gi$data$Patient <- as.character(p2gi$data$Patient)
p2gi$data$Patient <- factor(p2gi$data$Patient, levels=newPTorder)

# convert to interactive
ip1 <- ggplotly(p2gi + theme_bw())
ip1

# save as standalone html file
htmlwidgets::saveWidget(as_widget(ip1), "interactive.viraltax.taxplot.html")

```
# Viral Taxonomic 'Dotplot' by SampleType
```{r}
# TAXONOMIC 'DOTPLOT'

# Plot grouped by SampleType
# could consider coloring and grouping by higher level phylogeny

p3 <- group_by(ps2.sgdra.melt, SampleType, Species) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  na.omit() %>%
ggplot(aes(y=Species, x=SampleType)) +
  geom_point(aes(size = mean.abundance, color = Species)) +
  scale_color_viridis(discrete = TRUE, option = "plasma", guide = "none") +
  scale_size_continuous(range = c(5,30),
                        name = "Average Relative\nAbundance (%)") +
  ylab('Species') + xlab('Sample Type') +
  theme_bw() +
  theme(axis.text = element_text(size=20),
        axis.text.x = element_text(size = 22, angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank(),
        axis.title = element_text(size=24),
        legend.text = element_text(size=22),
        legend.position = "right",
        legend.title = element_text(size = 22, hjust = 0.5),
        plot.title = element_text(hjust = 0.5))+
  scale_y_discrete(limits=rev)
p3$data$SampleType <- as.character(p3$data$SampleType)
p3$data$SampleType <- factor(p3$data$SampleType, levels=newSTorder)
p3

# Save & print to file
png("viral.species.dotplot.draft.DEFINED.png", width = 12, height = 18, units = "in", res = 1000)
p3 + theme(text = element_text(size=15))
dev.off()

```

# For Irene's Revisions
```{r}
# Wound/Skin Avg TYPE Abundance
test.sp2 <- ps2.tgdra.melt %>%
  group_by(Description, Type) %>%
  filter(Abundance > 2) %>%
ggplot(aes(x= reorder(Type, +Abundance), y = Abundance, fill = Description)) +
  geom_point(stat='identity') +
  geom_boxplot() +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("FantasticFox1", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Type") + ylab("Relative Abundance (%)") +
  facet_wrap("Description", nrow=1)



test.sp2

# Print & save to file
png("sampletype.viraltype.boxplots.DESCRIP.DEFINED.png", width = 8, height = 8, units = "in", res = 1000)
test.sp2
dev.off()


# Wound/Skin Avg TYPE Abundance

test.sp2b <- ps2.sgdra.melt %>%
  group_by(Description, Species) %>%
  filter(Abundance > 5) %>%
ggplot(aes(x= reorder(Species, +Abundance), y = Abundance, fill = Description)) +
  geom_point(stat='identity') +
  geom_boxplot() +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("FantasticFox1", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Species") + ylab("Relative Abundance (%)") +
  facet_wrap("Description", nrow=1)



test.sp2b

# Print & save to file
png("sampletype.viraltype.boxplots.DESCRIP.DEFINED.png", width = 8, height = 8, units = "in", res = 1000)
test.sp2b
dev.off()
```



# host/type glom dotplot by sample type
```{r}
p3 <- group_by(ps2.tgdra.melt, SampleType, Type) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  na.omit() %>%
ggplot(aes(y=reorder(Type, desc(Type)), x=SampleType)) +
  geom_point(aes(size = mean.abundance, color = Type)) +
  scale_color_viridis(discrete = TRUE, option = "plasma", guide = "none") +
  scale_size_continuous(range = c(5,30),
                        name = "Average Relative\nAbundance (%)") +
  ylab('Type') + xlab('Sample Type') +
  theme_bw() +
  theme(axis.text = element_text(size=22),
        axis.text.x = element_text(size = 22, angle = 90, hjust = 1, vjust = 0.5),
        axis.title.x = element_blank(),
        axis.title = element_text(size=24),
        legend.text = element_text(size=22),
        legend.position = "right",
        legend.title = element_text(size = 22, hjust = 0.5),
        plot.title = element_text(hjust = 0.5))
p3$data$SampleType <- as.character(p3$data$SampleType)
p3$data$SampleType <- factor(p3$data$SampleType, levels=newSTorder)
p3

# Save & print to file
png("viral.type.dotplot.draft.DEFINED.png", width = 10.5, height = 16, units = "in", res = 600)
p3 + theme(text = element_text(size=15))
dev.off()


```

# Taxonomic 'Dotplot' and RA Boxplots by Outcome
```{r}
# Dot plot grouped by outcome
ps2.sgdra.melt %>% filter(SampleType != "Skin" & Outcome != "") %>% group_by(Outcome, Species) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  ggplot() +
  geom_point(aes(y=reorder(Species, desc(Species)), x=Outcome,
                 size = mean.abundance, color = Species)) +
  scale_size_continuous(range = c(5,30), name = "Average Relative\nAbundance (%)") +
  theme_classic() + ylab('Taxonomy') + xlab("Outcome") +
  theme(axis.text = element_text(size=18),
        axis.text.x = element_text(size = 18, angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_text(size=22),
        legend.text = element_text(size=18),
        legend.position = "none",
        legend.title = element_text(size = 22, hjust = 0.5),
        plot.title = element_text(hjust = 0.5)) -> wo1
wo1

png("viral.taxa.dotplot.outcome.DEFINED.png", width = 15, height = 12, units = "in", res = 600)
wo1
dev.off()

# boxplots

wo2 <- ps2.sgdra.melt %>%
  filter(SampleType != "Skin" & Outcome != "") %>%
  group_by(Outcome, Species) %>%
  filter(Abundance > 2) %>%
ggplot(aes(x= reorder(Species, +Abundance), y = Abundance, fill = Outcome)) +
  geom_point(stat='identity') +
  geom_boxplot() +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("Moonrise3", type = "discrete")) +
  theme_bw() +
  theme(strip.text.x = element_text(size = 18),
        axis.text = element_text(size=18),
        axis.text.x = element_text(size = 18, angle = 90, hjust = 1, vjust = 0.5),
        axis.title = element_text(size=22),
        legend.text = element_text(size=18),
        legend.position = "none",
        legend.title = element_text(size = 22, hjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  xlab("Taxonomy") + ylab("Relative Abundance (%)") +
  facet_wrap("Outcome", nrow=1)

wo2

# Print & save to file
png("outcome.viral.species.histos.DEFINED.png", width = 12, height = 12, units = "in", res = 600)
wo2
dev.off()

wo3 <- ps2.sgdra.melt %>%
  filter(SampleType != "Skin" & Outcome != "") %>%
  group_by(Outcome, SampleType, Species) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  ggplot(aes(x= reorder(Species, +mean.abundance), y = mean.abundance, fill = Outcome)) +
  geom_bar(stat='identity') +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("Moonrise3", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Taxonomy") + ylab("Average Relative Abundance (%)") +
  facet_wrap(Outcome ~ SampleType)
wo3

```
# Average RA Histograms by Sample Type
## Viral Species
```{r}
# AVERAGE RELATIVE ABUNDANCE HISTOS - SAMPLE TYPE, Viral Species
# Calculate average relative abundance of each species, plot
test.sp1 <- ps2.tpsgra.melt %>%
  group_by(SampleType, Type_PS) %>%
  summarise(mean.abundance = mean(Abundance, na.rm = T)) %>%
  filter(mean.abundance > 0.5) %>%
  ggplot(aes(x= reorder(Type_PS, +mean.abundance), y = mean.abundance, fill = SampleType)) +
  geom_bar(stat='identity') +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("FantasticFox1", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Type") + ylab("Average Relative Abundance (%)") +
  facet_wrap("SampleType", nrow=1)
test.sp1$data$SampleType <- as.character(test.sp1$data$SampleType)
test.sp1$data$SampleType <- factor(test.sp1$data$SampleType, levels=newSTorder)


test.sp1

# Print & save to file
png("avg.sampletype.vtypePS.histos.png", width = 6, height = 6.5, units = "in", res = 1200)
test.sp1
dev.off()

```
### Sample Type Boxplot - Species Level
```{r}
test.sp2 <- ps2.sgdra.melt %>%
  group_by(SampleType, Species) %>%
  filter(Abundance > 2) %>%
ggplot(aes(x= reorder(Species, +Abundance), y = Abundance, fill = SampleType)) +
  geom_point(stat='identity') +
  geom_boxplot() +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("FantasticFox1", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Species") + ylab("Relative Abundance (%)") +
  facet_wrap("SampleType", nrow=1)
test.sp2$data$SampleType <- as.character(test.sp2$data$SampleType)
test.sp2$data$SampleType <- factor(test.sp2$data$SampleType, levels=newSTorder)


test.sp2

# Print & save to file
png("sampletype.vspecies.boxplots.DEFINED.png", width = 8, height = 16, units = "in", res = 1000)
test.sp2
dev.off()
```
## Sample Type Boxplots - Type Level
```{r}
test.sp2 <- ps2.tgdra.melt %>%
  group_by(SampleType, Type) %>%
  filter(Abundance > 2) %>%
ggplot(aes(x= reorder(Type, +Abundance), y = Abundance, fill = SampleType)) +
  geom_point(stat='identity') +
  geom_boxplot() +
  coord_flip() + 
  scale_fill_manual(values = wes_palette("FantasticFox1", type = "discrete")) +
  theme_bw() +
  theme(legend.position="none") +
  xlab("Type") + ylab("Relative Abundance (%)") +
  facet_wrap("SampleType", nrow=1)
test.sp2$data$SampleType <- as.character(test.sp2$data$SampleType)
test.sp2$data$SampleType <- factor(test.sp2$data$SampleType, levels=newSTorder)


test.sp2

# Print & save to file
png("sampletype.viraltype.boxplots.DEFINED.png", width = 8, height = 8, units = "in", res = 1000)
test.sp2
dev.off()
```

# Top 300 OTU Heatmap
## Viral
```{r}
# TOP 300 OTU HEATMAP

# Prune to top 300 OTUs,  convert to relative abundance
hmf.ps2 <- ps2.pf.decon
hmf.ps2 <- prune_taxa(names(sort(taxa_sums(hmf.ps2),TRUE)[1:300]), hmf.ps2)
hmf.ps2.ra <- transform_sample_counts(hmf.ps2, function(x) (x / sum(x))*100 )
hmf.ps2.ra.melt <- psmelt(hmf.ps2.ra)
my_breaks = c(0.001, 0.01, 0.1, 1, 10, 50)
p4 = ggplot(hmf.ps2.ra.melt, aes(x = Patient, 
                                       y = reorder(OTU, Abundance))) +
              geom_tile(aes(fill = Abundance)) + 
              scale_fill_gradient(trans = "log",
                                  name = "Relative\nAbundance (%)",
                                  breaks = my_breaks,
                                  labels = my_breaks,
                                  na.value = "black") +
              facet_wrap("SampleType~.", nrow = 1, ncol = 3, 
                scales = "free") + ylab("Taxon") +
              theme(strip.text.x = element_text(size = 12),
                axis.text.y=element_blank(), 
                axis.ticks.y=element_blank(),
                panel.spacing.x = unit(0, "lines"),
                axis.title = element_text(size = 13),
                legend.title = element_text(size = 12),
                legend.text = element_text(size=12),
                axis.text.x = element_text(angle = 90, 
                                             hjust = 1, vjust = 0.5))
p4$data$SampleType <- as.character(p4$data$SampleType)
p4$data$SampleType <- factor(p4$data$SampleType, levels=newSTorder)
p4$data$Patient <- as.character(p4$data$Patient)
p4$data$Patient <- factor(p4$data$Patient, levels=newPTorder)
p4
# Print & save to file
png("viral.ra.heatmap.DEFINED.png", width = 8, height = 6, units = "in", res = 600)
p4
dev.off()
```
# Alpha Diversity Plots by Sample Type
## Viral
```{r}
# Plot
ps2.alpha <- subset_samples(ps2.pf.decon, SampleType %in% c("Pre-Debridement", "Post-Debridement", "Skin"))
p5 = plot_richness(ps2.alpha, x="SampleType", measures=c("Chao1", "Shannon"))

# # Fix X-axis order
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
p5$data$SampleType <- as.character(p5$data$SampleType)
p5$data$SampleType <- factor(p5$data$SampleType, levels=newSTorder)

# comparisons
alpha.comps = list(c("Pre-Debridement", "Post-Debridement"), c("Pre-Debridement", "Skin"), c("Post-Debridement", "Skin"))

# Plot with aesthetics and stats
alpha.div = p5 + theme_bw() + 
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust = 0.5),
        legend.position = "none") +
   stat_compare_means(method = "wilcox.test", alternative = "two-sided", 
                      comparisons = alpha.comps, paired = TRUE, 
                     aes(group = SampleType, label = ..p.adj..)) +
  #stat_compare_means(method = "wilcox.test", alternative = "two.sided",
  #                   comparisons = alpha.comps, paired = TRUE,
  #                   aes(group = SampleType, label = ..p.adj..)) +
  geom_line(aes(group=Patient), colour="grey", alpha = 0.5) + xlab("Sample Type")
                                                                   

#a.comp = compare_means(value ~ SampleType, alpha.div$data, method = "wilcox.test",
#                       paired = TRUE, group.by = "variable", alternative =
#                         "two.sided")
#a.comp
#library(rstatix)
#a.comp <- a.comp %>% get_y_position() %>% add_y_position(0.9)
#alpha.div = alpha.div + stat_pvalue_manual(a.comp, label = "p.adj")



# Print
ad = alpha.div + geom_point(aes(color = SampleType)) +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete")) 
ad.box = alpha.div + geom_boxplot(aes(fill = SampleType)) + scale_fill_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type")
ad
ad.box


# Save to file
png("alpha.div.draft.png", width = 5, height = 5, units = "in", res= 600)
alpha.div
dev.off()
png("alpha.div.draft.wbox.vir.png", width = 5, height = 5, units = "in", res= 600)
ad.box
dev.off()
```
### Get alpha div stats
```{r}
adstat <- ad.box$data

meanchao <- adstat %>% group_by(Description, variable) %>% summarise (meanchao = mean(value), sdchao = sd(value))

meanshan <- adstat %>% group_by(SampleType, variable) %>% filter(variable == "Shannon") %>% summarise (meanshan = mean(value), sdshan = sd(value))

meanchao
meanshan
```


# Alpha Diversity Plots by Outcome
```{r}
# ALPHA DIV - healed vs. unhealed

# Remove Skin samples
hvu.ss <- subset_samples(ps2.pf.decon, Outcome != "Skin")

# Plot
tp5 = plot_richness(hvu.ss, x="Outcome", color = "SampleType", measures=c("Chao1", "Shannon"))

# comparisons
alpha.comps = list(c("Healed", "Unhealed"))

# Plot with aesthetics and stats
hvu.alpha.div = tp5 + theme_bw() + 
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust = 0.5),
        text = element_text(size=15)) + 
  stat_compare_means(method = "wilcox.test", alternative = "two-sided", comparisons = alpha.comps, 
                     aes(group = Outcome, label = ..p.adj..)) +
  scale_color_manual(values = wes_palette("Darjeeling1", 2, type = "discrete"), name = "Sample Type")

# Print
hvu.alpha.div

# Save to file
png("alpha.div.OUTCOME.draft.png", width = 5, height = 5, units = "in", res= 600)
hvu.alpha.div
dev.off()
```
# Beta Diversity Ordinations
## Bray NMDS
```{r}
# BETA ORDINATIONS
ps2.p5 <- subset_samples(ps2.ra, Description == "Wound")
flist <- filterfun(kOverA(2,0.01))
ps2.p5 <- filter_taxa(ps2.p5, flist, TRUE)


# Ordination 1: NMDS BRAY ordination
ps2bray.ord <- ordinate(ps2.p5, "PCoA", "bray")
p6 = plot_ordination(ps2.p5, ps2bray.ord, type="samples", color="Outcome", shape = "SampleType") + 
  scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + 
  scale_shape(name = "Sample Type")
p6$data$SampleType <- as.character(p6$data$SampleType)
p6$data$SampleType <- factor(p6$data$SampleType, levels=newSTorder)
p6l <- p6 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Bray-Curtis PCoA") + theme_bw() + theme(legend.position = "right")
p6l

png("bray.pcoa.png", width = 6, height = 5, units = "in", res = 1200)
p6l
dev.off()
```

## Jaccard
```{r}
# Ordination 2: Jaccard
# try phylogenetic distance measures, start unweighted
ordu.uw = ordinate(ps2.p5, "PCoA", "jaccard")
p7 = plot_ordination(ps2.p5, ordu.uw, color="Outcome", shape = "SampleType") + scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + scale_shape(name = "Sample Type")
p7$data$SampleType <- as.character(p7$data$SampleType)
p7$data$SampleType <- factor(p7$data$SampleType, levels=newSTorder)
p7l <- p7 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Jaccard PCoA") + theme_bw()
p7l

```
## Manhattan
```{r}
# Ordination 3: Weighted UniFrac PCoA

# try phylogenetic distance measures, weighted now
ordu.w = ordinate(ps2.p5, "PCoA", "manhattan")
p8 = plot_ordination(ps2.p5, ordu.w, color="Outcome", shape = "SampleType") + scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + 
  scale_shape(name = "Sample Type")
p8$data$SampleType <- as.character(p8$data$SampleType)
p8$data$SampleType <- factor(p8$data$SampleType, levels=newSTorder)
p8l <- p8 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Manhattan PCoA") + theme_bw()
p8l
```
## Oridnation figure generation
```{r}
# horizontal grid print of all 3 ords
png("all.PCoA.horiz.OUTCOME.png", width = 15, height = 4, units = "in", res = 600)
grid.arrange(p6l, p7l, p8l, ncol=3)
dev.off()
```
# Beta Diversity Ordination by Outcome
```{r}
# ORD BY OUTCOME
# Bray
ps2bray.ord <- ordinate(ps2.p5, "PCoA", "bray")
p6 = plot_ordination(ps2.p5, ps2bray.ord, type="samples", color="Outcome", shape = "WoundType") + 
  scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + 
  scale_shape(name = "Wound Type")
outorder = c("Unhealed","Healed","Skin")
p6$data$Outcome <- as.character(p6$data$Outcome)
p6$data$Outcome <- factor(p6$data$Outcome, levels=outorder)
p6l <- p6 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Bray-Curtis PCoA") + theme_bw()
p6l

# Unweighted UniFrac
ordu.uw = ordinate(ps2.p5, "PCoA", "jaccard")
p7 = plot_ordination(ps2.p5, ordu.uw, color="Outcome", shape = "WoundType") + scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + scale_shape(name = "Wound Type")
p7$data$Outcome <- as.character(p7$data$Outcome)
p7$data$Outcome <- factor(p7$data$Outcome, levels=outorder)
p7l <- p7 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Jaccard PCoA") + theme_bw()
p7l

# Weighted UniFrac
ordu.w = ordinate(ps2.p5, "PCoA", "manhattan", weighted=TRUE)
p8 = plot_ordination(ps2.p5, ordu.w, color="Outcome", shape = "WoundType") + scale_color_manual(values = wes_palette("Darjeeling1", 3, type = "discrete"), name = "Outcome") + 
  scale_shape(name = "Wound Type")
p8$data$Outcome <- as.character(p8$data$Outcome)
p8$data$Outcome <- factor(p8$data$Outcome, levels=outorder)
p8l <- p8 + geom_line(aes(group=Pre.Post), colour="black") + geom_point(size=1.5) + ggtitle("Manhattan PCoA") + theme_bw()
p8l

# horizontal grid print of all 3 ords
png("all.PCoA.horiz.OUTCOME.png", width = 15, height = 4, units = "in", res = 600)
grid.arrange(p6l, p7l, p8l, ncol=3)
dev.off()
```
# Ordination Cluster Analysis

```{r}
# ORDINATION CLUSTER ANALYSIS
# calculate partitioning around medoids (PAM) and facet ordinations
library(cluster)
library(phyloseq)

BCdist = distance(ps2.pf, "bray")
bray.pcoa = ordinate(ps2.p5, "NMDS", BCdist)
plot_scree(bray.pcoa)
plot_ordination(ps2.p5, bray.pcoa, color = 'SampleType', shape = 'WoundType')

# BRAY-CURTIS
# bray hclust
hclustavg<-hclust(BCdist,method= "average")
plot(hclustavg, hang=-1, main="AverageLinkage")
rect.hclust(hclustavg,k =3)

# bray pam
pamClust3=pam(BCdist,k=3, cluster.only=T)
sample_data(ps2.p5)$pam3<-factor(pamClust3)
plot_ordination(ps2.p5, bray.pcoa, color ="SampleType")+
facet_wrap(~pam3, nrow =1)
p6l$data$pam3<-factor(pamClust3)
p6l
p6l + facet_wrap(~pam3, nrow =1)

# UNWEIGHTED UNIFRAC
# uw uni hclust
uw.dist = distance(ps2.p5, "unifrac")
hclustavg<-hclust(uw.dist, method= "average")
plot(hclustavg, hang=-1, main="AverageLinkage")
rect.hclust(hclustavg,k =4)

# uw uni pam
pamClust3=pam(uw.dist, k=4, cluster.only=T)
sample_data(ps2.p5)$pam3<-factor(pamClust3)
plot_ordination(ps2.p5, ordu.uw, color ="SampleType")+
facet_wrap(~pam3, nrow =1)
p7l$data$pam3<-factor(pamClust3)
p7l
p7l + facet_wrap(~pam3, nrow =2)

# WEIGHTED UNIFRAC
# w uni hclust
w.dist = distance(ps2.p5, "wunifrac")
hclustavg<-hclust(w.dist, method= "average")
plot(hclustavg, hang=-1, main="AverageLinkage")
rect.hclust(hclustavg,k =6)

# w uni pam
pamClust3=pam(w.dist,k=6, cluster.only=T)
sample_data(ps2.p5)$pam3<-factor(pamClust3)
plot_ordination(ps2.p5, ordu.w, color ="SampleType")+
facet_wrap(~pam3, nrow =1)
p8l$data$pam3<-factor(pamClust3)
p8l
p8l + facet_wrap(~pam3, nrow =2)
```

## Gap Statistics
```{r}
# CLUSTER ANALYSIS CONT'D
# get gap statistics for each
plot_scree(ps2bray.ord)
gs = gapstat_ord(ps2bray.ord, axes=c(1:2))
plot_clusgap(gs)
print(gs, method="Tibs2001SEmax")

plot_scree(ordu.uw)
gs = gapstat_ord(ordu.uw, axes=c(1:2))
plot_clusgap(gs)
print(gs, method="Tibs2001SEmax")

plot_scree(ordu.w)
gs = gapstat_ord(ordu.w, axes=c(1:3))
plot_clusgap(gs)
print(gs, method="Tibs2001SEmax")

```

## PerMANOVAs
```{r}
# CLUSTER(?) ANALYSIS CONT'D
# PerMANOVA tests across sample types (this tests correlation of multiple variables)
library(vegan)
#ent <-subset_samples(ent,!is.na(Enterotype))

df = data.frame(sample_data(ps2.p5))
BC = distance(ps2.p5, "bray")
adonis(BC~ Location + Outcome , data=df)
```

# Pairwise Beta Diversity Boxplots by Sample Type
```{r}
# Pairewise Beta diversity plots (separated by metric)
plotDistances = function(p = ps2.p5, b = "bray", u = "jaccard", w = "manhattan", s = "SampleID", d = "SampleType", g = "Patient") {

  # calc weighted unifrac distances
  bc = phyloseq::distance(p, b)
  bc.m = melt(as.matrix(bc))
  # calc manhattan distances
  wu = phyloseq::distance(p, w)
  wu.m = melt(as.matrix(wu))
  # calc jaccard distances
  uw = phyloseq::distance(p, u)
  uw.m = melt(as.matrix(uw))
  
  # remove self-comparisons
    # bray
    bc.m = bc.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
    # manhattan
    wu.m = wu.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
    # jaccard
    uw.m = uw.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
  
  # get sample data (S4 error OK and expected)
  sd = sample_data(p) %>%
    select(s, d, g) %>%
    mutate_if(is.factor,as.character)
  
  # combined distances with sample data
    # Var1
    colnames(sd) = c("Var1", "Type1", "Patient1")
      # bray
      bc.sd = left_join(bc.m, sd, by = "Var1")
      # manhattan
      wu.sd = left_join(wu.m, sd, by = "Var1")
      # jaccard
      uw.sd = left_join(uw.m, sd, by = "Var1")
    # Var2
    colnames(sd) = c("Var2", "Type2", "Patient2")
      # bray
      bc.sd = left_join(bc.sd, sd, by = "Var2")
      # manhattan
      wu.sd = left_join(wu.sd, sd, by = "Var2")
      # jaccard
      uw.sd = left_join(uw.sd, sd, by = "Var2")
  
  # Filter to get only within-patient comparisons
  # bray
  bc.sd = bc.sd %>%
    filter(as.character(Patient1) == as.character(Patient2)) %>%
    mutate_if(is.factor,as.character)
  # manhattan
  wu.sd = wu.sd %>%
    filter(as.character(Patient1) == as.character(Patient2)) %>%
    mutate_if(is.factor,as.character)
  # jaccard
  uw.sd = uw.sd %>%
    filter(as.character(Patient1) == as.character(Patient2)) %>%
    mutate_if(is.factor,as.character)
  
  # Label comparison types
  # bray
  bc.sd$comps = 
    ifelse(bc.sd$Type1 == "Pre-Debridement" & 
           bc.sd$Type2 == "Post-Debridement", "Pre vs. Post", 
    ifelse(bc.sd$Type1 == "Pre-Debridement" & 
           bc.sd$Type2 == "Skin", "Pre vs. Skin",
    ifelse(bc.sd$Type1 == "Post-Debridement" & 
           bc.sd$Type2 == "Skin", "Post vs. Skin", NA)))
  bc.sd = bc.sd %>% filter(as.character(comps) != "NA")
  
  # manhattan
  wu.sd$comps = 
    ifelse(wu.sd$Type1 == "Pre-Debridement" & 
           wu.sd$Type2 == "Post-Debridement", "Pre vs. Post", 
    ifelse(wu.sd$Type1 == "Pre-Debridement" & 
           wu.sd$Type2 == "Skin", "Pre vs. Skin",
    ifelse(wu.sd$Type1 == "Post-Debridement" & 
           wu.sd$Type2 == "Skin", "Post vs. Skin", NA)))
  wu.sd = wu.sd %>% filter(as.character(comps) != "NA")
  
  # jaccard
  uw.sd$comps = 
    ifelse(uw.sd$Type1 == "Pre-Debridement" &
           uw.sd$Type2 == "Post-Debridement", "Pre vs. Post",
    ifelse(uw.sd$Type1 == "Pre-Debridement" &
           uw.sd$Type2 == "Skin", "Pre vs. Skin",
    ifelse(uw.sd$Type1 == "Post-Debridement" &
           uw.sd$Type2 == "Skin", "Post vs. Skin", NA)))
  uw.sd = uw.sd %>% filter(as.character(comps) != "NA")
  
  # plot
bcp1 <- ggplot(bc.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point() +
    geom_line(aes(group=Patient1), colour="grey", alpha = 0.5) +
    geom_boxplot(aes(fill=comps)) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "none") + 
    ggtitle("Bray-Curtis Distance")
  
wup1 <- ggplot(wu.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point() +
    geom_line(aes(group=Patient1), colour="grey", alpha = 0.5) +
    geom_boxplot(aes(fill=comps)) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "none") + 
    ggtitle("Manhattan")

uwp1 <- ggplot(uw.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point() +
    geom_line(aes(group=Patient1), colour="grey", alpha = 0.5) +
    geom_boxplot(aes(fill=comps)) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "none") + 
    ggtitle("Jaccard")

# fix sample order
newCorder = c("Pre vs. Skin","Pre vs. Post","Post vs. Skin")
bcp1$data$comps <- as.character(bcp1$data$comps)
bcp1$data$comps <- factor(bcp1$data$comps, levels=newCorder)
wup1$data$comps <- as.character(wup1$data$comps)
wup1$data$comps <- factor(wup1$data$comps, levels=newCorder)
uwp1$data$comps <- as.character(uwp1$data$comps)
uwp1$data$comps <- factor(uwp1$data$comps, levels=newCorder)

# define stat comparisons
beta.comps = list(c("Pre vs. Post", "Pre vs. Skin"), c("Pre vs. Post", "Post vs. Skin"), c("Pre vs. Skin", "Post vs. Skin"))
beta.groups = c(u,w)

# print with stats
bcps = bcp1 + 
  stat_compare_means(method = "wilcox.test", alternative = "two.sided", comparisons = beta.comps, paired = TRUE, step.increase = 0.15, aes(label = ..p.adj..)) + 
  ylab("Distance") + ylim(0,1.4) +
  xlab("Comparison") + 
  scale_fill_manual(values = wes_palette("Moonrise2", 3, type = "discrete"))
wups = wup1 + 
  stat_compare_means(method = "wilcox.test", alternative = "two.sided", comparisons = beta.comps, paired = TRUE, step.increase = 0.15, aes(label = ..p.adj..)) + 
  ylab("Distance") + ylim(0,275) +
  xlab("Comparison") + 
  scale_fill_manual(values = wes_palette("Moonrise2", 3, type = "discrete"))
uwps = uwp1 + 
  stat_compare_means(method = "wilcox.test", alternative = "two.sided", comparisons = beta.comps, paired = TRUE, step.increase = 0.15, aes(label = ..p.adj..)) + 
  ylab("Distance") + ylim(0,1.4) +
  xlab("Comparison") +
  scale_fill_manual(values = wes_palette("Moonrise2", 3, type = "discrete"))
grid.arrange(bcps, uwps, wups, ncol = 3)

# collect mean & sd for each plot
b.mean <- group_by(bcp1$data, comps) %>% summarise(b.mean = mean(value))
b.sd <- group_by(bcp1$data, comps) %>% summarise(b.sd = sd(value))
b.stat = left_join(b.mean, b.sd)

w.mean <- group_by(wup1$data, comps) %>% summarise(w.mean = mean(value))
w.sd <- group_by(wup1$data, comps) %>% summarise(w.sd = sd(value))
w.stat = left_join(w.mean, w.sd)

u.mean <- group_by(uwp1$data, comps) %>% summarise(u.mean = mean(value))
u.sd <- group_by(uwp1$data, comps) %>% summarise(u.sd = sd(value))
u.stat = left_join(u.mean, u.sd)

uw.stat = left_join(u.stat, w.stat)
uwb.stat = left_join(uw.stat, b.stat)

}
plotDistances()


png("beta.diversity.boxplots.DEFINED.png", width = 11, height = 5, units = "in", res = 1200)
plotDistances()
dev.off()
```

# Beta Diversity Boxplots - Wounds vs. Skin by Outcome
```{r}
# BETA DIV BOXPLOTS - Wound vs. Skin by Outcome

plotDistances = function(p = ps2.p5, u = "bray", w = "jaccard", s = "SampleID", d = "SampleType", g = "Patient", o = "Outcome") {

  # calc weighted unifrac distances
  wu = phyloseq::distance(p, w)
  wu.m = melt(as.matrix(wu))
  # calc unweighted unifrac distances
  uw = phyloseq::distance(p, u)
  uw.m = melt(as.matrix(uw))
  
  # remove self-comparisons
    # weighted
    wu.m = wu.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
    # unweighted
    uw.m = uw.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
  
  # get sample data (S4 error OK and expected)
  sd = sample_data(p) %>%
    select(s, d, g, o) %>%
    mutate_if(is.factor,as.character)
  
  # combined distances with sample data
    # Var1
    colnames(sd) = c("Var1", "Type1", "Patient1", "Outcome1")
      # weighted
      wu.sd = left_join(wu.m, sd, by = "Var1")
      # unweighted
      uw.sd = left_join(uw.m, sd, by = "Var1")
    # Var2
    colnames(sd) = c("Var2", "Type2", "Patient2", "Outcome2")
      # weighted
      wu.sd = left_join(wu.sd, sd, by = "Var2")
      # unweighted
      uw.sd = left_join(uw.sd, sd, by = "Var2")
  
  # Filter to get only within-patient comparisons
  # weighted
  wu.sd = wu.sd %>%
    filter(as.character(Patient1) == as.character(Patient2)) %>%
    mutate_if(is.factor,as.character)
  # unweighted
  uw.sd = uw.sd %>%
    filter(as.character(Patient1) == as.character(Patient2)) %>%
    mutate_if(is.factor,as.character)
  
  # Filter to remove NA outcomes
  wu.sd = wu.sd %>%
    filter(as.character(Outcome1) != "") %>% mutate_if(is.factor,as.character)
  uw.sd = uw.sd %>%
    filter(as.character(Outcome1) != "") %>% mutate_if(is.factor,as.character)
  
  # Label comparison types
  # weighted
  wu.sd$comps = 
    ifelse(wu.sd$Type1 == "Pre-Debridement" & 
           wu.sd$Type2 == "Post-Debridement", "Pre vs. Post", 
    ifelse(wu.sd$Type1 == "Pre-Debridement" & 
           wu.sd$Type2 == "Skin", "Pre vs. Skin",
    ifelse(wu.sd$Type1 == "Post-Debridement" & 
           wu.sd$Type2 == "Skin", "Post vs. Skin", NA)))
  wu.sd = wu.sd %>% filter(as.character(comps) != "NA")
  
  # unweighted
  uw.sd$comps = 
    ifelse(uw.sd$Type1 == "Pre-Debridement" &
           uw.sd$Type2 == "Post-Debridement", "Pre vs. Post",
    ifelse(uw.sd$Type1 == "Pre-Debridement" &
           uw.sd$Type2 == "Skin", "Pre vs. Skin",
    ifelse(uw.sd$Type1 == "Post-Debridement" &
           uw.sd$Type2 == "Skin", "Post vs. Skin", NA)))
  uw.sd = uw.sd %>% filter(as.character(comps) != "NA")
  
  # plot
wup1 <- ggplot(wu.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point(position=position_dodge(width=0.75), aes(group = Outcome1)) +
    geom_boxplot(aes(fill=Outcome1)) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "bottom") + 
    ggtitle("Jaccard")

uwp1 <- ggplot(uw.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point(position=position_dodge(width=0.75), aes(group = Outcome1)) +
    geom_boxplot(aes(fill=Outcome1)) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "bottom") +
    ggtitle("Bray")

# fix sample order
newCorder = c("Pre vs. Skin","Pre vs. Post","Post vs. Skin")
wup1$data$comps <- as.character(wup1$data$comps)
wup1$data$comps <- factor(wup1$data$comps, levels=newCorder)
uwp1$data$comps <- as.character(uwp1$data$comps)
uwp1$data$comps <- factor(uwp1$data$comps, levels=newCorder)

# define stat comparisons
#beta.comps = list("Healed","Unhealed")
#beta.groups = c(u,w)

# print with stats
wups = wup1 + 
  stat_compare_means(aes(group=Outcome1), method = "wilcox.test", alternative = "two.sided", label = "p.signif", label.y.npc = 1) +
  ylab("Distance") +
  xlab("Comparison") + ylim(0,1.2) +
  scale_fill_manual(values = wes_palette("Royal1", 3, type = "discrete"), name = "Outcome") + 
  theme(text = element_text(size=15))

uwps = uwp1 + 
  stat_compare_means(aes(group=Outcome1), method = "wilcox.test", alternative = "two.sided", label = "p.signif", label.y.npc = 1) +
  ylab("Distance") +
  xlab("Comparison") + ylim(0,1.2) +
  scale_fill_manual(values = wes_palette("Royal1", 3, type = "discrete"), name = "Outcome") +
  theme(text = element_text(size=15))

grid.arrange(uwps, wups, ncol = 2)

}

plotDistances()

png("bd.by.outcome.boxplots.draft.png", width = 8, height = 5, units = "in", res = 600)
plotDistances()
dev.off()
```


# PAIRWISE BETA BY OUTCOME
```{r}
# PAIRWISE BETA BY OUTCOME
## Need to compare distance b/w healed/unhealed wounds
## do dist within healed and unhealed, and the distance between them

bd.out <- subset_samples(ps2.p5, SampleType == "Pre-Debridement")

plotDistances = function(p = bd.out, u = "bray", w = "jaccard", s = "SampleID", g = "Patient", o = "Outcome") {

  # calc jaccard distances
  wu = phyloseq::distance(p, w)
  wu.m = melt(as.matrix(wu))
  # calc bray distances
  uw = phyloseq::distance(p, u)
  uw.m = melt(as.matrix(uw))
  
  # remove self-comparisons
    # jaccard
    wu.m = wu.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
    # bray
    uw.m = uw.m %>%
      filter(as.character(Var1) != as.character(Var2)) %>%
      mutate_if(is.factor,as.character)
  
  # get sample data (S4 error OK and expected)
  sd = sample_data(p) %>%
    select(s, g, o) %>%
    mutate_if(is.factor,as.character)
  
  # combined distances with sample data
    # Var1
    colnames(sd) = c("Var1", "Patient1", "Outcome1")
      # jaccard
      wu.sd = left_join(wu.m, sd, by = "Var1")
      # bray
      uw.sd = left_join(uw.m, sd, by = "Var1")
    # Var2
    colnames(sd) = c("Var2", "Patient2", "Outcome2")
      # jaccard
      wu.sd = left_join(wu.sd, sd, by = "Var2")
      # bray
      uw.sd = left_join(uw.sd, sd, by = "Var2")
  

  # Filter to remove NA outcomes
  wu.sd = wu.sd %>%
    filter(as.character(Outcome1) != "") %>% mutate_if(is.factor,as.character)
  uw.sd = uw.sd %>%
    filter(as.character(Outcome1) != "") %>% mutate_if(is.factor,as.character)
  
  # Label comparison types
  # jaccard
  wu.sd$comps = 
    ifelse(wu.sd$Outcome1 == "Healed" & 
           wu.sd$Outcome2 == "Unhealed", "Healed vs. Unhealed", 
    ifelse(wu.sd$Outcome1 == "Healed" & 
           wu.sd$Outcome2 == "Healed", "Healed vs. Healed",
    ifelse(wu.sd$Outcome1 == "Unhealed" & 
           wu.sd$Outcome2 == "Unhealed", "Unhealed vs. Unhealed", NA)))
  wu.sd = wu.sd %>% filter(as.character(comps) != "NA")
  
  # bray
  uw.sd$comps = 
    ifelse(uw.sd$Outcome1 == "Healed" & 
           uw.sd$Outcome2 == "Unhealed", "Healed vs. Unhealed", 
    ifelse(uw.sd$Outcome1 == "Healed" & 
           uw.sd$Outcome2 == "Healed", "Healed vs. Healed",
    ifelse(uw.sd$Outcome1 == "Unhealed" & 
           uw.sd$Outcome2 == "Unhealed", "Unhealed vs. Unhealed", NA)))
  uw.sd = uw.sd %>% filter(as.character(comps) != "NA")
  
  # plot
wup1 <- ggplot(wu.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point(position=position_dodge(width=0.75)) +
    geom_boxplot() +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "bottom") + 
    ggtitle("Jaccard")

uwp1 <- ggplot(uw.sd, aes(x = comps, y = value)) +
    theme_bw() +
    geom_point(position=position_dodge(width=0.75)) +
    geom_boxplot(aes()) +
    theme(axis.text.x=element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "bottom") +
    ggtitle("Bray-Curtis")

# fix sample order
newCorder = c("Healed vs. Healed","Healed vs. Unhealed","Unhealed vs. Unhealed")
wup1$data$comps <- as.character(wup1$data$comps)
wup1$data$comps <- factor(wup1$data$comps, levels=newCorder)
uwp1$data$comps <- as.character(uwp1$data$comps)
uwp1$data$comps <- factor(uwp1$data$comps, levels=newCorder)

# define stat comparisons
# comparisons
beta.comps = list(c("Healed vs. Healed", "Healed vs. Unhealed"), c("Healed vs. Healed", "Unhealed vs. Unhealed"), c("Healed vs. Unhealed", "Unhealed vs. Unhealed"))

# print with stats
wups = wup1 + 
  stat_compare_means(method = "wilcox.test", paired = FALSE, comparisons = beta.comps, step.increase = 0.3, aes(label = ..p.adj..)) +
  ylab("Distance") +
  xlab("Comparison") + ylim(0,1.5) +
  #scale_fill_manual(values = wes_palette("Royal1", 3, type = "discrete"), name = "Outcome") + 
  theme(text = element_text(size=15))

uwps = uwp1 + 
  stat_compare_means(method = "wilcox.test", paired = FALSE, comparisons = beta.comps, step.increase = 0.25, aes(label = ..p.adj..)) +
  ylab("Distance") +
  xlab("Comparison") + ylim(0,1.5) +
  #scale_fill_manual(values = wes_palette("Royal1", 3, type = "discrete"), name = "Outcome") +
  theme(text = element_text(size=15))

grid.arrange(uwps, wups, ncol = 2)

}

plotDistances()

png("bd.by.outcome.boxplots.draft.png", width = 8, height = 5, units = "in", res = 600)
plotDistances()
dev.off()
```


## Positive Control Only: Expected vs. Observed
```{r}
# Enter mock community info (from Zymo)
mock.genera = c("Pseudomonas", "Escherichia-Shigella", "Salmonella", "Lactobacillus", "Enterococcus", "Staphylococcus", "Listeria", "Bacillus")
RAbundance = c(4.2, 10.1, 10.4, 18.4, 9.9, 15.5, 14.1, 17.4)
# build into dataframe
pos.comp <- data.frame(mock.genera, RAbundance)
colnames(pos.comp)[1] <- "Genus"
pos.comp$Sample[1:8] <- "Expected"

# prepare microbial standard data
## Subset
poscon <- subset_samples(ra.tf.ps2.con, SampleType == "Microbial_Standard")
## melt
posmelt <- psmelt(poscon)
## average the samples and filter to include only taxa in mock community
pos.avg <- posmelt %>% group_by(Genus) %>% summarise(RAbundance = (mean(Abundance))*100) %>% filter(Genus %in% mock.genera)
pos.avg$Sample[1:8] <- "Observed"

# combine tables
pos.comp = rbind(pos.comp, pos.avg)
pos.comp

# plot
pcp1 <- pos.comp %>% 
  group_by(Sample, Genus) %>%
  ggplot(aes(x=Sample, y=RAbundance, fill = Genus)) +
  geom_bar(stat="identity") +
  theme_bw() + ylab("Average Relative Abundance (%)") + 
  theme(legend.position = "right", axis.title.x = element_blank()) +
  scale_fill_brewer(palette = "Set1", name = "Genus")
pcp1

png("positive.control.comp.png", width = 5, height = 5, units = "in", res = 600)
pcp1
dev.off()

```

# Viral relabund vs Bacterial Relabund
```{r}
### make separate plot for staphylococcus species and pseudomonas species?

# 16S IMPORT
# Import biom table and mapping
biom_file <- "silva.sorted.raw.biom"
map_file <- "ps2mapping.16S.dryad.txt"
biom_otu_tax <- import_biom(biom_file, parseFunction=parse_taxonomy_default)

# Rename taxonomic levels
colnames(tax_table(biom_otu_tax)) <- c(Rank1="Domain",Rank2="Phylum",Rank3="Class",Rank4="Order",Rank5="Family",Rank6="Genus",Rank7="Species")

# Add sample data to the dataset w/ merge
bmsd <- import_qiime_sample_data(map_file)
dim(bmsd)
ps2.16s <- merge_phyloseq(biom_otu_tax, bmsd)
ps2.16s

#  Remove the SILVA prefix "D_#__" from the taxonomy
tax_table(ps2.16s) <- gsub("D_[0-9]__","", tax_table(ps2.16s))

# Sample filtering and agglomeration
ps2.16s.cor <- subset_samples(ps2.16s, SampleType %in% c("Pre-Debridement","Post-Debridement","Skin"))
ps2.16s.cor.gg <- tax_glom(ps2.16s.cor, "Genus", NArm = FALSE)
```


```{r}
## Agglommerate at the Species and Type levels
### Sample objects
ps2.sgd.ss <- subset_taxa(ps2.pf.decon, SResolution == "Defined")
ps2.sgd <- tax_glom(ps2.sgd.ss, "Species", NArm = FALSE)
ps2.tgd.ss <- subset_taxa(ps2.pf.decon, TResolution == "Defined")
ps2.tgd <- tax_glom(ps2.tgd.ss, "Type", NArm = FALSE)
ps2.sgdra <- transform_sample_counts(ps2.sgd, function(x) (x / sum(x))*100 )
ps2.tgdra <- transform_sample_counts(ps2.tgd, function(x) (x / sum(x))*100 )
```


# Phage:Host Correlation
```{r}
# Viral type abundance data
typecor = as.data.frame(sample_data(ps2.tgdra))
typecor <- typecor %>% select("SampleID", "SampleType", "Pre.Post", "Outcome")
typecor$pseudovir =  sample_sums(subset_taxa(ps2.tgdra, Type == "Pseudomonas phage"))
typecor$staphvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Staphylococcus phage"))
typecor$strepvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Streptococcus phage"))
typecor$proteusvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Proteus phage"))
typecor$bacteroidesvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Bacteroides phage"))
typecor$corynevir = sample_sums(subset_taxa(ps2.tgdra, Type == "Corynebacterium phage"))
typecor$enterovir = sample_sums(subset_taxa(ps2.tgdra, Type %in% c("Enterobacter phage", "Enterobacteriaceae phage", "Enterobacteria phage", "Enterococcus phage")))
typecor$actinovir = sample_sums(subset_taxa(ps2.tgdra, Type %in% c("Actinobaculum phage", "Actinobacteria phage", "Propionibacterium phage", "Corynebacterium phage", "Micrococcus phage", "Streptomyces phage", "Actinomyces phage")))
typecor$propiovir = sample_sums(subset_taxa(ps2.tgdra, Type == "Propionibacterium phage"))
typecor$neissvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Neisseria phage"))
typecor$campyvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Campylobacter phage"))
typecor$bacillvir = sample_sums(subset_taxa(ps2.tgdra, Type == "Bacillus phage"))
typecor
```

## add abundances to correlation dataframe
```{r}
bac.ra <- transform_sample_counts(ps2.16s.cor.gg, function(x) x / sum(x)*100 )
typecor$pseudobac =  sample_sums(subset_taxa(bac.ra, Genus == "Pseudomonas"))
typecor$staphbac = sample_sums(subset_taxa(bac.ra, Genus == "Staphylococcus"))
typecor$strepbac = sample_sums(subset_taxa(bac.ra, Genus == "Streptococcus"))
typecor$proteusbac = sample_sums(subset_taxa(bac.ra, Genus == "Proteus"))
typecor$bacteroidesbac = sample_sums(subset_taxa(bac.ra, Genus == "Bacteroides"))
typecor$corynebac = sample_sums(subset_taxa(bac.ra, Genus == "Corynebacterium 1"))
typecor$enterobac = sample_sums(subset_taxa(bac.ra, Family == "Enterobacteriaceae"))
typecor$actinobac = sample_sums(subset_taxa(bac.ra, Phylum == "Actinobacteria"))
typecor$propiobac = sample_sums(subset_taxa(bac.ra, Genus == "Propionibacterium"))
typecor$neissbac = sample_sums(subset_taxa(bac.ra, Genus == "Neisseria"))
typecor$campybac = sample_sums(subset_taxa(bac.ra, Genus == "Campylobacter"))
typecor$bacillbac = sample_sums(subset_taxa(bac.ra, Genus == "Bacillus"))
typecor
```
## Prep table
```{r}
# problems transposing, export and transpose outside of R then reimport
#typecor.t <- as.data.frame(t(typecor))
#typecor.t
#write.csv(typecor.t, file = "bac.vir.cor.transposed.csv")

# vOTU import
biom_file <- "bac.vir.cor.transposed.csv"
otu.csv <- read.csv(biom_file, row.names = 1, header = TRUE, check.names = FALSE)
ps2.otu.mat <- as.matrix(otu.csv, rownames=TRUE)
mode(ps2.otu.mat) <- "numeric"

# Build phylo object
#bmsd <- import_qiime_sample_data(map_file)
OTU <- otu_table(ps2.otu.mat, taxa_are_rows = TRUE)
SAM <- sample_data(ps2.pf)
ps2.bvcor <- merge_phyloseq(OTU, SAM)
ps2.bvcor

# import and add taxonomy
tax_file <- "bac.vir.cor.tax.csv"
ps2.cortax <- read.csv(tax_file, row.names = 1, header = TRUE, check.names = FALSE, na.strings=c("","NA"))
ps2.cortax_table <- as.matrix(ps2.cortax)
tax_table(ps2.bvcor) <- ps2.cortax_table
ps2.bvcor
ps2.bvcor.melt <- psmelt(ps2.bvcor)
ps2.bvcor.melt


cordf <- ps2.bvcor.melt %>% filter(bac_or_vir == "vir") %>% select("Sample", "SampleType", "Pre.Post", "Outcome")
cordf

vircor.ab <- ps2.bvcor.melt %>% filter(bac_or_vir == "vir") %>% select("Sample", "Abundance", "tax_group")
colnames(vircor.ab)[2] <- "VirAbund"
vircor.ab

baccor.ab <- ps2.bvcor.melt %>% filter(bac_or_vir == "bac") %>% select("Sample", "Abundance", "tax_group")
colnames(baccor.ab)[2] <- "BacAbund"
baccor.ab

cordf <- merge(x = cordf, y = vircor.ab, by.x = "Sample", by.y = "Sample")
cordf <- merge(x = cordf, y = baccor.ab, by.x = c("Sample", "tax_group"), by.y = c("Sample","tax_group"))
cordf <- unique(cordf)
cordf
```
## plot
```{r}
bvcorplots <- ggplot(cordf, aes(x = VirAbund, y = BacAbund, color = SampleType, shape = Outcome)) + 
  geom_line(aes(group=Pre.Post), color="black") +
  geom_point() + facet_wrap("tax_group") + 
  xlab("Viral relative abundance (%)") +
  ylab("Bacterial relative abundance (%)") +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type") +
  theme_bw() #+
  #scale_x_log10() + scale_y_log10() + theme(axis.text.x=element_text(angle = 90, hjust = 0, vjust = 0.5))
  
bvcorplots

bvcorscat <- ggscatter(cordf, x="VirAbund", y="BacAbund", color = "SampleType", shape = "Outcome") + stat_cor(method = "pearson", label.x.npc = 0.15, show.legend = FALSE) +
  geom_line(aes(group=Pre.Post), color="black") + 
  facet_wrap("tax_group") + 
  theme_bw() + 
  xlab("Viral relative abundance (%)") +
  ylab("Bacterial relative abundance (%)") +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type") +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        strip.text.x = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14))

bvcorscat$data$SampleType <- as.character(bvcorscat$data$SampleType)
bvcorscat$data$SampleType <- factor(bvcorscat$data$SampleType, levels=newSTorder)
bvcorscat
#png("all.bacterial.viral.cor.LOG.png", width = 8, height = 6, units = "in", res = 1000)
#bvcorplots
#dev.off()

png("all.bacterial.viral.cor.stat.png", width = 12, height = 10, units = "in", res = 600)
bvcorscat
dev.off()
```

# Alpha div: Bacterial vs. Viral
```{r}
# bacterial alpha
bac.alpha <- subset_samples(ps2.16s.cor, SampleType %in% c("Pre-Debridement", "Post-Debridement", "Skin"))
bacalpha.plot = plot_richness(bac.alpha, x="SampleType", measures=c("Shannon"))
bacadivdat <- data.frame(bacalpha.plot$data)
bacadivdat

# viral
vir.alpha <- subset_samples(ps2.pf.decon, SampleType %in% c("Pre-Debridement", "Post-Debridement", "Skin"))
viralpha.plot = plot_richness(vir.alpha, x="SampleType", measures=c("Shannon"))
viradivdat <- data.frame(viralpha.plot$data)
viradivdat

combalpha <- data.frame(bacalpha.plot$data$SampleID)
colnames(combalpha) <- "SampleID"
combalpha$bac <- bacadivdat$value
combalpha$vir <- viradivdat$value
combalpha$SampleType <- viradivdat$SampleType
combalpha$Outcome <- viradivdat$Outcome
combalpha$Pre.Post <- viradivdat$Pre.Post
combalpha

alphacor <- ggplot(combalpha, aes(x = bac, y = vir, color = SampleType, shape = Outcome)) + 
  geom_line(aes(group=Pre.Post), colour = "black") + geom_point(size=2) + 
  xlab("Bacterial Shannon Index") + ylab("Viral Shannon Index") +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type") +
  scale_shape(name = "Outcome") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box = 'horizontal',
        legend.direction = 'vertical',
        #legend.justification = 'left',
        legend.box.just = 'left') +
        stat_cor(inherit.aes = FALSE, aes(x = bac, y = vir), method = "spearman", label.x.npc = 0.5)
  

alphacor$data$SampleType <- as.character(alphacor$data$SampleType)
alphacor$data$SampleType <- factor(alphacor$data$SampleType, levels=newSTorder)
alphacor


alphacor.stat <- ggscatter(combalpha, x = "bac", y = "vir", color = "SampleType", shape = "Outcome") +
  stat_cor(inherit.aes = FALSE, 
           aes(x = bac, y = vir), method = "pearson", 
           add = "reg.line", conf.int = TRUE) + 
  geom_line(aes(group=Pre.Post), colour = "black") +
  xlab("Bacterial Shannon Index") + ylab("Viral Shannon Index") +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type") +
  scale_shape(name = "Outcome") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box = 'horizontal',
        legend.direction = 'vertical',
        #legend.justification = 'left',
        legend.box.just = 'left')
alphacor.stat$data$SampleType <- as.character(alphacor.stat$data$SampleType)
alphacor.stat$data$SampleType <- factor(alphacor.stat$data$SampleType, levels=newSTorder)
alphacor.stat


alpheat = ggplot(hmf.ps2.ra.melt, aes(x = Patient, 
                                       y = reorder(OTU, +Abundance))) +
              geom_tile(aes(fill = Abundance)) + 
              scale_fill_gradient(trans = "log",
                                  name = "Relative\nAbundance (%)",
                                  breaks = my_breaks,
                                  labels = my_breaks,
                                  na.value = "black") +
              facet_wrap("SampleType~.", nrow = 1, ncol = 3, 
                scales = "free") + ylab("Taxon") + theme_bw() +
              theme(#strip.text.x = element_text(size = 12),
                axis.text.x=element_blank(), 
                axis.ticks.x=element_blank(),
                axis.text.y=element_blank(), 
                axis.ticks.y=element_blank(),
                legend.position = "bottom", legend.key.width = unit(1.25,"cm"))

#png("alpha.div.correlation.png", width = 6, height = 5, units = "in", res = 1000)
#alphacor
#dev.off()
```


```{r}
#png("full.alphadiv.fig.png", width = 12, height = 6, units = "in", res = 1000)
#plot_grid(ad.box, alphacor, ncol = 2, align = "h", axis = "b", rel_widths = c(1, 0.85))
#dev.off()
admod = ad.box + theme(axis.title.x = element_blank())

#png("full.alphadiv.fig.png", width = 4, height = 10, units = "in", res = 1000)
#plot_grid(admod, alphacor, ncol = 1, align = "v", axis = "l")
#dev.off()

png("full.alphadiv.fig.png", width = 12, height = 6, units = "in", res = 1000)
plot_grid(admod, alpheat, alphacor, ncol = 3, nrow = 1, align = "h", axis = "b", rel_widths = c(1, 1, 1))
dev.off()
```


# Bacteria-Bacteriophage Co-occurrence & Correlation
# Bacterial genus vs. Phage type
```{r}

```



# viral and unclassified counts
```{r}
ps2.dg.vu <- subset_taxa(ps2.dg, Domain %in% c("Viruses","unclassified"))
ps2.dg.vu

# make a boxplot, sort by patient, stack pre-, post-, skin
p2d = plot_bar(ps2.dg.vu, fill="Domain", facet_grid = "SampleType~.", x="Patient") #+ scale_fill_manual(values = genus.colors)
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
# facet order
newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
p2d$data$SampleType <- as.character(p2d$data$SampleType)
p2d$data$SampleType <- factor(p2d$data$SampleType, levels=newSTorder)
# x-axis order
newPTorder = as.character(c(1:20))
p2d$data$Patient <- as.character(p2d$data$Patient)
p2d$data$Patient <- factor(p2d$data$Patient, levels=newPTorder)

p2df = p2d + theme_bw() + 
        theme(legend.position = "bottom",
        text = element_text(size=20),
        strip.text.y = element_text(size=24),
        axis.title = element_text(size=26),
        axis.text = element_text(size=22),
        legend.title = element_text(size=24)) + scale_y_log10() +
  guides(fill = guide_legend(title = "Domain", nrow=2))
p2df

# Print & save to file
png("vir.unclass.cts.taxplot.wlegend.png", width = 14, height = 14, units = "in", res = 600)
p2df
dev.off()

```

# Rarefaction / alpha vs. abundance or conc.
```{r}
# rarefaction 
library(vegan)
#ps4rare <- subset_samples(ps2.pf.decon, Patient  != "7")
#rarecurve(t(otu_table(ps4rare)), step=10)

out <- rarecurve(t(otu_table(ps2.pf.decon)), step = 10, label = FALSE)
```

```{r}
names(out) <- paste("species", 1:50, sep = "")

# Coerce data into "long" form.
protox <- mapply(FUN = function(x, y) {
  mydf <- as.data.frame(x)
  colnames(mydf) <- "value"
  mydf$species <- y
  mydf$subsample <- attr(x, "Subsample")
  mydf
}, x = out, y = as.list(names(out)), SIMPLIFY = FALSE)

xy <- do.call(rbind, protox)
rownames(xy) <- NULL  # pretty

# Plot.
ggplot(xy, aes(x = subsample, y = value, color = species)) +
  theme_bw() +
  scale_color_discrete(guide = FALSE) +  # turn legend on or off
  geom_line() + scale_x_log10()



```
## plot alpha against library size
```{r}
# alpha diversity values
viradivdat

# library sizes
adivls <- data.frame(sample_sums(ps2.pf))
colnames(adivls)[1] <- "libsize"
adivls$SampleID <- rownames(adivls)
adivls

# combine
adiv.vs.ls <- merge(viradivdat, adivls, by.x = 1, by.y = 2)
adiv.vs.ls

# plot
adls.p1 <- ggplot(adiv.vs.ls, aes(x = libsize, y = value, color = SampleType, shape = Outcome)) + 
  geom_line(aes(group=Pre.Post), colour = "black") + geom_point(size=2) + 
  xlab("Viral library size") + ylab("Viral Shannon Index") +
  scale_color_manual(values = wes_palette("FantasticFox1", 3, type = "discrete"), name = "Sample Type") +
  scale_shape(name = "Outcome") +
  theme_bw() + scale_x_log10() +
  theme(legend.position = "bottom", 
        legend.box = 'horizontal',
        legend.direction = 'vertical',
        #legend.justification = 'left',
        legend.box.just = 'left') +
        stat_cor(inherit.aes = FALSE, aes(x = libsize, y = value), method = "pearson", label.x.npc = 0.7)
adls.p1$data$SampleType <- as.character(adls.p1$data$SampleType)
adls.p1$data$SampleType <- factor(adls.p1$data$SampleType, levels=newSTorder)
adls.p1

```


# Spearman Correlations
```{r}
library(microbiome)
library(data.table)

# bacterial filter
bac.ra.ss <- subset_samples(bac.ra, Description == "Wound")
flist <- filterfun(kOverA(2,1))
ps2.16s.precorr <- filter_taxa(bac.ra.ss, flist, TRUE)
bac.ra.ss
ps2.16s.precorr

# viral filter
ps2.sgdra.ss <- subset_samples(ps2.sgdra, Description == "Wound")
flist <- filterfun(kOverA(2,1))
ps2v.pf.precorr <- filter_taxa(ps2.sgdra.ss, flist, TRUE)
ps2.sgdra.ss
ps2v.pf.precorr

# prepare the bacterial dataframe
# fix this so the genus names are the column names?
bdf <- data.frame(otu_table(ps2.16s.precorr), check.names = FALSE)
bdf.tax <- data.frame(ps2.16s.precorr@tax_table@.Data, check.names = FALSE)
bdf.taxnona <- bdf.tax %>% 
  rownames_to_column("id") %>% 
  filter(Genus != "NA") %>%
  column_to_rownames("id")
bdf.nona <- bdf %>% 
  rownames_to_column("id") %>%
  filter(rownames(bdf) %in% rownames(bdf.taxnona)) %>%
  column_to_rownames("id")
row.names(bdf.nona) <- bdf.taxnona$Genus
bdf.nona
bdf.t <- transpose(bdf.nona)
colnames(bdf.t) <- rownames(bdf.nona)
rownames(bdf.t) <- colnames(bdf.nona)
bdf
bdf.t

# prepare the viral dataframe
# fix this so the species names are the column names?
vdf <- data.frame(otu_table(ps2v.pf.precorr), check.names = FALSE)
vdf.tax <- data.frame(ps2v.pf.precorr@tax_table@.Data, check.names = FALSE)
vdf.taxnona <-vdf.tax %>% 
  rownames_to_column("id") %>% 
  filter(Species != "NA") %>%
  column_to_rownames("id")
vdf.nona <- vdf %>% 
  rownames_to_column("id") %>%
  filter(rownames(vdf) %in% rownames(vdf.taxnona)) %>%
  column_to_rownames("id")
row.names(vdf.nona) <- vdf.taxnona$Species
vdf.nona
vdf.t <- transpose(vdf)
colnames(vdf.t) <- rownames(vdf.nona)
rownames(vdf.t) <- colnames(vdf.nona)
vdf
vdf.t

```

## associate
```{r}
cc <- associate(vdf.t, bdf.t, method='spearman', order = TRUE, mode = 'table', filter.self.correlations = TRUE)
cc
cc.m <- associate(vdf.t, bdf.t, method='spearman', order = TRUE, mode = 'matrix', filter.self.correlations = TRUE)
cc.m


## Fix the matrix
# preview the correlation table
cc
cc.mt.full <- as.data.frame(cc.m)
cc.mt.full

cc.mt <- cc.mt.full[1:27]
names(cc.mt) <- substring(names(cc.mt), 5)

cc.mm <- data.matrix(cc.mt, rownames.force = TRUE)
class(cc.mm)

#heatmap
library(pheatmap)
paletteLength <- 1000
myColor <- colorRampPalette(c("blue", "white", "red"))(paletteLength)
# length(breaks) == length(paletteLength) + 1
# use floor and ceiling to deal with even/odd length pallettelengths
myBreaks <- c(seq(min(cc.mm), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(cc.mm)/paletteLength, max(cc.mm), length.out=floor(paletteLength/2)))

pheatmap(cc.mm, col=myColor, breaks=myBreaks, cutree_rows = 5, angle_col = "45", filename="wound.heatmap.png", width = 10, height = 14)


pd.m <- melt( cc.mt, id.vars = "Virus", variable.name = "Bacteria" )
pd.m

# add taxonomy
# prep bacterial genus
bcortax.pre <- as.data.frame(ps2.16s.precorr@tax_table@.Data)
bcortax <- data.frame(bcortax.pre$Genus)
colnames(bcortax)[1] <- "Genus"
bcortax

# prep viral species
vcortax.pre <- as.data.frame(ps2v.pf.precorr@tax_table@.Data)
vcortax <- data.frame(vcortax.pre$Species)
colnames(vcortax)[1] <- "Species"
vcortax

# merge
ccc.tax <- merge(pd.m, bcortax, by.x = 2, by.y = 0)
ccc.tax <- merge(ccc.tax, vcortax, by.x = 2, by.y = 0)
ccc.tax
```


## add taxonomy to cor table
```{r}
# prep bacterial genus
bcortax.pre <- as.data.frame(ps2.16s.precorr@tax_table@.Data)
bcortax <- data.frame(bcortax.pre$Genus)
colnames(bcortax)[1] <- "Genus"
bcortax

# prep viral species
vcortax.pre <- as.data.frame(ps2v.pf.precorr@tax_table@.Data)
vcortax <- data.frame(vcortax.pre$Species)
colnames(vcortax)[1] <- "Species"
vcortax

# merge
cc.tax <- merge(cc, bcortax, by.x = 2, by.y = 0)
cc.tax <- merge(cc.tax, vcortax, by.x = 2, by.y = 0)
cc.tax
```

## Plot as heatmap
```{r}
cc.temp <- cc.tax %>% filter(p.adj <= 0.05) #%>% filter(abs(Correlation) >= 0.5)
cc.temp


## FULL
pal <- wes_palette("Zissou1", 100, type = "continuous")
corrheat <- cc.tax %>% filter(Genus != "NA") %>%
  ggplot(aes(x = reorder(Species, +Correlation), 
                       y = reorder(Genus, +Correlation), 
                       fill = Correlation)) + 
  geom_tile() +
  #scale_fill_gradientn(colours = pal) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(fill = "Spearman\ncorrelation\ncoefficient") +
  xlab("Viral species") + ylab("Bacterial genus")
corrheat

png("heat.corr.OVERALL.FULL.png", width = 14, height = 8, units = "in", res = 600)
corrheat
dev.off()

## SIGNIFICANT
corrheat <- cc.temp %>% filter(Genus != "NA") %>%
  ggplot(aes(x = reorder(Species, +Correlation), 
                       y = reorder(Genus, +Correlation), 
                       fill = Correlation)) + 
  geom_tile() +
  #scale_fill_gradientn(colours = pal) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(fill = "Spearman\ncorrelation\ncoefficient") +
  xlab("Viral species") + ylab("Bacterial genus")
corrheat

png("heat.corr.OVERALL.SIG.png", width = 10, height = 8, units = "in", res = 600)
corrheat
dev.off()
```


## Repeat for wounds only
```{r}
# bacterial filter
bac.ra.ss <- subset_samples(bac.ra, Description == "Wound")
flist <- filterfun(kOverA(2,1))
ps2.16s.precorr <- filter_taxa(bac.ra.ss, flist, TRUE)
bac.ra.ss
ps2.16s.precorr

# viral filter
ps2.sgdra.ss <- subset_samples(ps2.sgdra, Description == "Wound")
flist <- filterfun(kOverA(2,1))
ps2v.pf.precorr <- filter_taxa(ps2.sgdra.ss, flist, TRUE)
ps2.sgdra.ss
ps2v.pf.precorr

# prepare the bacterial dataframe
# fix this so the genus names are the column names?
bdf <- data.frame(otu_table(ps2.16s.precorr), check.names = FALSE)
bdf.tax <- 
bdf.t <- transpose(bdf)
colnames(bdf.t) <- rownames(bdf)
rownames(bdf.t) <- colnames(bdf)
bdf
bdf.t

# prepare the viral dataframe
# fix this so the species names are the column names?

vdf <- data.frame(otu_table(ps2v.pf.precorr), check.names = FALSE)
vdf.t <- transpose(vdf)
colnames(vdf.t) <- rownames(vdf)
rownames(vdf.t) <- colnames(vdf)
vdf
vdf.t
```
## wound associate
```{r}
cc <- associate(vdf.t, bdf.t, method='spearman', order = TRUE, mode = 'table', filter.self.correlations = TRUE)
cc
```
## prep tax and plot
```{r}
# preview the correlation table
cc

# prep bacterial genus
bcortax.pre <- as.data.frame(ps2.16s.precorr@tax_table@.Data)
bcortax <- data.frame(bcortax.pre$Genus)
colnames(bcortax)[1] <- "Genus"
bcortax

# prep viral species
vcortax.pre <- as.data.frame(ps2v.pf.precorr@tax_table@.Data)
vcortax <- data.frame(vcortax.pre$Species)
colnames(vcortax)[1] <- "Species"
vcortax

# merge
cc.tax <- merge(cc, bcortax, by.x = 2, by.y = 0)
cc.tax <- merge(cc.tax, vcortax, by.x = 2, by.y = 0)
cc.tax

cc.temp <- cc.tax %>% filter(p.adj <= 0.05) #%>% filter(abs(Correlation) >= 0.5)
#cc.temp

## FULL
pal <- wes_palette("Zissou1", 100, type = "continuous")
corrheat.w <- cc.tax %>% filter(Genus != "NA") %>%
            ggplot(aes(x = reorder(Species, +Correlation), 
                       y = reorder(Genus, +Correlation), 
                       fill = Correlation)) + 
  geom_tile() +
  #scale_fill_gradientn(colours = pal) + 
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(fill = "Spearman\ncorrelation\ncoefficient") +
  xlab("Viral species") + ylab("Bacterial genus") #+ ggtitle("Wound Correlations")
corrheat.w

png("heat.corr.WOUNDS.FULL.png", width = 12, height = 6, units = "in", res = 600)
corrheat.w
dev.off()

## SIGNIFICANT
pal <- wes_palette("Zissou1", 100, type = "continuous")
corrheat.w <- cc.temp %>% filter(Genus != "NA") %>%
            ggplot(aes(x = Species, 
                       y = Genus, 
                       fill = Correlation)) + 
  geom_tile() +
  #scale_fill_gradientn(colours = pal) + 
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(fill = "Spearman\ncorrelation\ncoefficient") +
  xlab("Viral species") + ylab("Bacterial genus") #+ ggtitle("Wound Correlations")
corrheat.w

png("heat.corr.WOUNDS.SIG.png", width = 6, height = 4, units = "in", res = 600)
corrheat.w
dev.off()
```

## Repeat for skin
```{r}
# bacterial filter
bac.ra.ss <- subset_samples(bac.ra, Description == "Skin")
flist <- filterfun(kOverA(2,0.1))
ps2.16s.precorr <- filter_taxa(bac.ra.ss, flist, TRUE)
bac.ra.ss
ps2.16s.precorr

# viral filter
ps2.sgdra.ss <- subset_samples(ps2.sgdra, Description == "Skin")
flist <- filterfun(kOverA(2,0.1))
ps2v.pf.precorr <- filter_taxa(ps2.sgdra.ss, flist, TRUE)
ps2.sgdra.ss
ps2v.pf.precorr

# prepare the bacterial dataframe
# fix this so the genus names are the column names?
bdf <- data.frame(otu_table(ps2.16s.precorr), check.names = FALSE)
bdf.tax <- 
bdf.t <- transpose(bdf)
colnames(bdf.t) <- rownames(bdf)
rownames(bdf.t) <- colnames(bdf)
bdf
bdf.t

# prepare the viral dataframe
# fix this so the species names are the column names?

vdf <- data.frame(otu_table(ps2v.pf.precorr), check.names = FALSE)
vdf.t <- transpose(vdf)
colnames(vdf.t) <- rownames(vdf)
rownames(vdf.t) <- colnames(vdf)
vdf
vdf.t

```

## skin associate
```{r}
cc <- associate(vdf.t, bdf.t, method='spearman', order = TRUE, mode = 'table', filter.self.correlations = TRUE)
cc
```
## prep tax and plot
```{r}
# preview the correlation table
cc

# prep bacterial genus
bcortax.pre <- as.data.frame(ps2.16s.precorr@tax_table@.Data)
bcortax <- data.frame(bcortax.pre$Genus)
colnames(bcortax)[1] <- "Genus"
bcortax

# prep viral species
vcortax.pre <- as.data.frame(ps2v.pf.precorr@tax_table@.Data)
vcortax <- data.frame(vcortax.pre$Species)
colnames(vcortax)[1] <- "Species"
vcortax

# merge
cc.tax <- merge(cc, bcortax, by.x = 2, by.y = 0)
cc.tax <- merge(cc.tax, vcortax, by.x = 2, by.y = 0)
cc.tax

cc.temp <- cc.tax %>% filter(p.adj <= 0.05) #%>% filter(abs(Correlation) >= 0.5)
#cc.temp

pal <- wes_palette("Zissou1", 5, type = "continuous")
corrheat.s <- ggplot(cc.tax, 
                   aes(x = reorder(Species, +Correlation), 
                       y = reorder(Genus, +Correlation), 
                       fill = Correlation)) + 
  geom_tile() +
  #scale_fill_gradientn(colours = pal) + 
  scale_fill_gradientn(colours = c("blue", "white", "red")) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  xlab("Viral species") + ylab("Bacterial genus") + ggtitle("Skin Correlations")
corrheat.s

png("heat.corr.SKIN.ALL.png", width = 8, height = 8, units = "in", res = 600)
corrheat.s
dev.off()
```


# Differential Abundance Analysis w/ DESeq2
```{r}
library(phyloseq)
library(genefilter)
library(ggplot2)
library(ggpubr)
library(reshape2)
library(RColorBrewer)
library(viridis)
library(wesanderson)
library(grid)
library(gridExtra)
library(plotly)
library(scales)
library(dplyr)
set.seed(1)
```


# File Import & Phyloseq Conversion
```{r}
# Import mapping
map_file <- "/Users/sverbanic/Desktop/PS2VB.taxtables/ps2mapping.deseq.virome.txt"
deseq.map <- import_qiime_sample_data(map_file)

SAM <- sample_data(deseq.map)

# use phyloseq object "ps2.sgd" generated in virome.primary.analysis.Rmd workup
ps2v.pf <- ps2.sgd
sample_data(ps2v.pf) <- SAM
```

# DESeq Analyses
```{r}
# DESeq2 differential abundance analysis
# Overall comparisons

library(DESeq2)

flist <- filterfun(kOverA(2,10))
ps2v.pf <- filter_taxa(ps2v.pf, flist, TRUE)
otu_table(ps2v.pf) <- otu_table(ps2v.pf) + 1

# These lines will produce the original DESeq2 object, and run the first comparison 
# convert to DESeq2
ps2.dds = phyloseq_to_deseq2(ps2v.pf, ~ 1)
# refactor levels for desired comparisons
ps2.dds$SampleType <- factor(ps2.dds$SampleType, levels = c("Pre-Debridement","Post-Debridement","Skin"))
# specify experiment design
design(ps2.dds) <- ~ SampleType
# run DESeq
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")

# Pre vs. Skin Results
res.ps = results(dds.out, contrast = c("SampleType","Pre-Debridement","Skin"), cooksCutoff = FALSE)
write.csv(res.ps, "pre.vs.skin.csv")

# Pre vs. Post Overall Results
res.pp = results(dds.out, contrast = c("SampleType","Pre-Debridement","Post-Debridement"), cooksCutoff = FALSE)
write.csv(res.pp, "pre.vs.post.overall.csv")

# Wound (Pre + Post) vs. Skin DESeq & Results
ps2.dds$Description <- as.factor(ps2.dds$Description)
design(ps2.dds) <- ~ Description
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.ws = results(dds.out, contrast = c("Description","Wound","Skin"), cooksCutoff = FALSE)
write.csv(res.ws, "wound.vs.skin.overall.csv")

# Pre vs. Post HEALED ONLY DESeq & Results
# Create nested group & run DESeq
ps2.dds$group <- factor(paste0(ps2.dds$Outcome, ps2.dds$prepost))
design(ps2.dds) <- ~ group
dds.hu = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.hpp = results(dds.hu, contrast = c("group","HealedPre","HealedPost"), cooksCutoff = FALSE)
write.csv(res.hpp, "pre.vs.post.healed.csv")

# Healed vs. Unhealed (Post- Only) Results
res.hupost = results(dds.hu, contrast = c("group","HealedPost","UnhealedPost"), cooksCutoff = FALSE)
write.csv(res.hupost, "healed.vs.unhealed.post.csv")

# Pre vs. Post UNHEALED ONLY DESeq & Results
# Create nested group & run DESeq
ps2.dds$group <- factor(paste0(ps2.dds$Outcome, ps2.dds$prepost))
# Refactor to get desired comparisons
ps2.dds$group <- factor(ps2.dds$group, levels = c("UnhealedPre","UnhealedPost","HealedPre","HealedPost","SkinControl"))
design(ps2.dds) <- ~ group
dds.hu = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.upp = results(dds.hu, contrast = c("group","UnhealedPre","UnhealedPost"), cooksCutoff = FALSE)
write.csv(res.upp, "pre.vs.post.unhealed.csv")

# Healed vs. Unhealed (Skin) DESeq & Results
ps2.dds$SkinOutcome <- as.factor(ps2.dds$SkinOutcome)
design(ps2.dds) <- ~ SkinOutcome
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.hu.skin = results(dds.out, contrast = c("SkinOutcome","Healed","Unhealed"), cooksCutoff = FALSE)
write.csv(res.hu.skin, "healed.vs.unhealed.skin.csv")

# Healed vs. Unhealed (Overall) DESeq & Results
ps2.dds$Outcome <- as.factor(ps2.dds$Outcome)
design(ps2.dds) <- ~ Outcome
dds.out = DESeq(ps2.dds, test="Wald", fitType="parametric")
res.hu = results(dds.out, contrast = c("Outcome","Healed","Unhealed"), cooksCutoff = FALSE)
write.csv(res.hu, "healed.vs.unhealed.overall.csv")

# Healed vs. Unhealed (Pre- Only) Results
res.hupre = results(dds.hu, contrast = c("group","HealedPre","UnhealedPre"), cooksCutoff = FALSE)
write.csv(res.hupre, "healed.vs.unhealed.pre.csv")
```


# Significant OTU Visualizations
## Pre vs. Skin
```{r}
# estimate 95% confidence intervals
res.ps$error = (res.ps$lfcSE)
res.ps

pal <- wes_palette("Zissou1", 3, type = "continuous")

alpha = 0.05
sigtab = res.ps[which(res.ps$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2v.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) + geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.5, position=position_dodge(0.05), colour = "black") + geom_point(size=6) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 22), legend.position = "none", plot.title = element_text(hjust = 0.5)) + 
  ggtitle("Skin vs. Pre-Debridement") + ylab("Species")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1

sigrows = row.names(as.data.frame(p1$data))
outpal <- wes_palettes$Zissou1

p1b <- ps2.sgdra.melt %>% filter(ps2.sgdra.melt$OTU %in% sigrows)  %>%
  filter(SampleType != "Post-Debridement") %>%
  filter(!Abundance <= 0.0001) %>%
  group_by(Outcome, Species) %>%
  ggplot(aes(y=Species, x=Abundance, color=SampleType)) +
  geom_point(size=4) + theme_bw() +  
  theme(text = element_text(size = 22), 
        legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("Relative abundance") + 
  scale_color_manual(values = outpal[c(5,1)], name = "Sample Type") +
  scale_x_log10() + xlab("Relative abundance (%)")
p1b  

x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
p1b$data$Species = factor(as.character(p1b$data$Species), levels=names(x))

png("pre.vs.skin.species.png", width = 20, height = 18, units = "in", res = 600)
grid.arrange(p1, p1b, widths=c(1,0.9), ncol = 2)
dev.off()
```

## Wounds (Pre + Post) vs. Skin
```{r}
# calc error
res.ws$error = (1.96*res.ws$lfcSE)
res.ws

alpha = 0.05
sigtab = res.ws[which(res.ws$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), width=0.5, height=0.5, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 22), legend.position = "none") + 
  ggtitle("Skin vs. Wound (overall)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1



sigrows = row.names(as.data.frame(p1$data))
outpal <- wes_palettes$Zissou1

p1b <- ps2.sgdra.melt %>% filter(ps2.sgdra.melt$OTU %in% sigrows)  %>%
  filter(!Abundance <= 0.0001) %>%
  group_by(Outcome, Species) %>%
  ggplot(aes(y=Species, x=Abundance, color=Description)) +
  geom_point(size=4) + theme_bw() +  
  theme(text = element_text(size = 22), 
        legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Relative abundance") + 
  scale_color_manual(values = outpal[c(1,5)], name = "Sample Type") +
  scale_x_log10() + xlab("Relative abundance (%)")
p1b  

x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
p1b$data$Species = factor(as.character(p1b$data$Species), levels=names(x))

png("wound.vs.skin.overall.png", width = 24, height = 16, units = "in", res = 600)
grid.arrange(p1, p1b, ncol = 2)
dev.off()
```


## Pre vs. Post (Overall)
```{r}
# Pre vs. Post Overall

# calculate error
res.pp$error = (1.96*res.pp$lfcSE)
res.pp

alpha = 0.01
sigtab = res.pp[which(res.pp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Species), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), width=0.5, height=0.25, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 22), legend.position = "none") + 
  ggtitle("Post- vs. Pre- (overall)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1


sigrows = row.names(as.data.frame(p1$data))
outpal <- wes_palettes$Zissou1

p1b <- ps2.sgdra.melt %>% filter(ps2.sgdra.melt$OTU %in% sigrows)  %>%
  filter(SampleType != "Skin") %>%
  filter(!Abundance <= 0.0001) %>%
  group_by(Outcome, Species) %>%
  ggplot(aes(y=Species, x=Abundance, color=SampleType)) +
  geom_point(size=4) + theme_bw() +  
  theme(text = element_text(size = 22), 
        legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Relative abundance") + 
  scale_color_manual(values = outpal[c(1,5)], name = "Sample Type") +
  scale_x_log10() + xlab("Relative abundance (%)")
p1b  

x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
p1b$data$Species = factor(as.character(p1b$data$Species), levels=names(x))

png("pre.vs.post.overall.TYPE.DECONNED.png", width = 16, height = 16, units = "in", res = 600)
grid.arrange(p1, p1b, ncol = 2)
dev.off()
```
## Pre vs. Post (Healed Only)
```{r}
# calc error
res.hpp$error = (1.96*res.hpp$lfcSE)
res.hpp

alpha = 0.05
sigtab = res.hpp[which(res.hpp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"),
as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.25, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 22), legend.position = "none") + 
  ggtitle("Post vs. Pre (Healed Only)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1


sigrows = row.names(as.data.frame(p1$data))
outpal <- wes_palettes$Zissou1

p1b <- ps2.sgdra.melt %>% filter(ps2.sgdra.melt$OTU %in% sigrows)  %>%
  filter(SampleType != "Skin") %>%
  filter(!Abundance <= 0.000001) %>%
  group_by(Outcome, Species) %>%
  ggplot(aes(y=Species, x=Abundance, color=SampleType)) +
  geom_point(size=4) + theme_bw() +  
  theme(text = element_text(size = 22), 
        legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Relative abundance") + 
  scale_color_manual(values = outpal[c(5,1)], name = "Sample Type") +
  scale_x_log10() + xlab("Relative abundance (%)")
p1b  

newSTorder = c("Pre-Debridement","Post-Debridement","Skin")
# color order
newSTorder = c("Pre-Debridement","Post-Debridement")
p1b$data$SampleType <- as.character(p1b$data$SampleType)
p1b$data$SampleType <- factor(p1b$data$SampleType, levels=newSTorder)

png("pre.vs.post.healed.png", width = 16, height = 6, units = "in", res = 600)
grid.arrange(p1, p1b, ncol = 2)
dev.off()
```
## Pre vs. Post (Unhealed Only)
```{r}
# calc error
res.upp$error = (1.96*res.upp$lfcSE)
res.upp

alpha = 0.05
sigtab = res.upp[which(res.upp$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot

p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.1, position=position_dodge(0.05), colour = "black") + 
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 18)) + ggtitle("Post vs. Pre (Unhealed Only)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("pre.vs.post.unhealed.png", width = 12, height = 12, units = "in", res = 600)
p1
dev.off()
```
## Healed vs. Unhealed (Overall)
```{r}
# Healed vs. Unhealed Overall
pal <- wes_palette("Zissou1", 3, type = "continuous")

# calculate error
res.hu$error = (res.hu$lfcSE)
res.hu

# alpha filter
alpha = 0.01
sigtab = res.hu[which(res.hu$padj < alpha), ]
# add taxonomy
sigtab = cbind(as(sigtab, "data.frame"),
as(tax_table(ps2v.pf)[rownames(sigtab), ], "matrix"))

# Set Y order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot log2fold change
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=-log2FoldChange)) + geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.25, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + 
  theme(text = element_text(size = 22), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) + 
  ggtitle("Unhealed vs. Healed") + ylab("Species")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1


sigrows = row.names(as.data.frame(p1$data))
outpal <- wes_palettes$Zissou1

p1b <- ps2.sgdra.melt %>% filter(ps2.sgdra.melt$OTU %in% sigrows)  %>%
  filter(Outcome != "Skin") %>%
  filter(!Abundance <= 0.0001) %>%
  group_by(Outcome, Species) %>%
  ggplot(aes(y=Species, x=Abundance, color=Outcome)) +
  geom_point(size=4) + theme_bw() +  
  theme(text = element_text(size = 22), 
        legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
        ggtitle("Species Relative Abundance") +
  scale_color_manual(values = outpal[c(2,5)], name = "Outcome") +
  scale_x_log10() + xlab("Relative abundance (%)")
p1b  

x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
p1b$data$Species = factor(as.character(p1b$data$Species), levels=names(x))

png("healed.vs.unhealed.overall.species.png", width = 20, height = 14, units = "in", res = 1000)
grid.arrange(p1, p1b, widths=c(1,0.75), ncol = 2)
dev.off()

p1$data
```
# Healed vs Unhealed SKIN
```{r}
# calculate error

res.hu.skin$error = (1.96*res.hu.skin$lfcSE)

alpha = 0.05
sigtab = res.hu.skin[which(res.hu.skin$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"),
as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.25, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + ggtitle("Unhealed vs. Healed (Skin Only)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.skin.png", width = 6, height = 6, units = "in", res = 600)
p1
dev.off()
```
## Healed vs. Unhealed (Pre- Only)
```{r}
# Healed vs. Unhealed Pre- only

# calculate error
res.hupre$error = (1.96*res.hupre$lfcSE)

alpha = 0.05
sigtab = res.hupre[which(res.hupre$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"),
as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) +
  geom_errorbarh(aes(xmin=log2FoldChange-error, xmax=log2FoldChange+error), height=0.25, position=position_dodge(0.05), colour = "black") +
  geom_point(size=8) + theme_bw() + scale_color_gradientn(colours = pal) + theme(text = element_text(size = 22)) + ggtitle("Unhealed vs. Healed (Pre- Only)")
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.pre.png", width = 12, height = 12, units = "in", res = 600)
p1
dev.off()
```

## Healed vs Unhealed (Post- Only)
```{r}
# Healed vs. Unhealed Post- only
alpha = 0.05
sigtab = res.hupost[which(res.hupost$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"),
as(tax_table(ps2.pf)[rownames(sigtab), ], "matrix"))
head(sigtab)
dim(sigtab)
x = tapply(sigtab$log2FoldChange, sigtab$Type, function(x) max(x))
x = sort(x, TRUE)
sigtab$Type = factor(as.character(sigtab$Type), levels=names(x))

# Species order
x = tapply(sigtab$log2FoldChange, sigtab$Species, function(x) max(x))
x = sort(x, TRUE)
sigtab$Species = factor(as.character(sigtab$Species), levels=names(x))

# plot
# 
p1 <- ggplot(sigtab, aes(y=Species, x=log2FoldChange, color=log2FoldChange)) + 
  geom_point(size=4) + theme_bw() + scale_color_gradientn(colours = pal)
  #theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
p1
png("healed.vs.unhealed.post.png", width = 6, height = 12, units = "in", res = 600)
p1
dev.off()
```

